---
title: "Proportional Groundfish Redistribution"
author: "Mason Smith (mason.smith@noaa.gov)"
date: today
format: 
  html:
    toc: true
    code-fold: true
    self-contained: true
params:
  time_step: "weekly" # Options: "weekly", "monthly", "annual"
  timeframe: [2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024, 2025]
  psc_spp: RKCR
  closure: rkcssa
  buffer: 100000
  gear: NPT
  sector: CP
  management_code: 
  catch_type: GF_TOTAL_CATCH_MT   #GF_TOTAL_CATCH_MT, SHALLOW_FLATFISH_WEIGHT
  lme: bs # (goa, ai, bs, or all)
  method: both #options: m1, m2, both
execute:
  warning: false
  message: false

---

## 1. Setup

This analysis predicts the spatial movement and net PSC change for displaced vessels under closures of the red king crab sub savings area (RKCSSA) and validates predictions against closures in 2022 and 2023. 

```{r}
#| label: setup
#| include: false

# Restrict multi-threading memory spikes in OpenBLAS
Sys.setenv(OPENBLAS_NUM_THREADS = "1")
Sys.setenv(OMP_NUM_THREADS = "1")
Sys.setenv(MKL_NUM_THREADS = "1")

# Load packages
library(dplyr)
library(terra)
library(ggplot2)
library(lubridate)
library(tidyr)
library(units)
library(gridExtra)
library(tidyterra)
library(ggpubr)
library(patchwork)
library(ggpattern)
library(data.table)
library(sf)

options(scipen=999)

```


## 2. Set Titles

Assign titles, timeframe, and closures

```{r}

spp_title = case_when(
      params$psc_spp == "OTCR"  ~ "Snow Crab",
      params$psc_spp == "BTCR"  ~ "Tanner Crab",
      params$psc_spp == "RKCR"  ~ "Red King Crab",
      params$psc_spp == "HERR"  ~ "Herring",
      TRUE                          ~ NA # Default: NA string
    )


title_name = case_when(
  params$closure == "smihca"         ~ "Saint Matthews HCA",
  params$closure == "nunivak"        ~ "Nunivak HCA",
  params$closure == "rkcssa"         ~ "Red King Crab Sub Savings Area",
  params$closure == "Alt1El1Opt2_2025" ~ "GOA Custom Closure Area", # <--- ADD THIS
  TRUE                               ~ NA_character_
)


```



## 3. Spatial Data and Grid Setup

Loading the Alaska background and statistical area shapefiles to create the time-series grid.

```{r}

#| label: load-spatial
stat_areas <- terra::vect("shapefiles/State_Stat_2004.shp")

# Ensure it stays a character string and strip any accidental whitespace
stat_areas$STAT_AREA <- trimws(as.character(stat_areas$STAT_AREA))

#filter to lme
# Define NMFS reporting area vector dynamically
target_areas <- switch(
  tolower(params$lme),
  "goa" = c(610, 620, 630, 640, 650, 649, 659),
  "ai"  = c(541, 542, 543),
  "bs"  = c(518, 519, 517, 513, 509, 516, 512, 508, 514, 400, 524, 521, 523, 550, 530),
  "all" = c(610, 620, 630, 640, 650, 649, 659,
            541, 542, 543,
            518, 519, 517, 513, 509, 516, 512, 508, 514, 400, 524, 521, 523, 550, 530),
  stop("Invalid params$lme value! Must be 'goa', 'ai', 'bs', or 'all'.")
)

# Then filter your stat_areas object cleanly
stat_areas <- stat_areas %>%
  dplyr::filter(GNMFSAREA %in% target_areas)

# Create timeseries list
stat.list <- list()
for (i in params$timeframe){
  stat_shape <- stat_areas
  stat_shape$YEAR <- i
  stat.list[[as.character(i)]] <- stat_shape
}
stat.areas.timeseries <- terra::vect(stat.list)


```

## 4. Closure Assignment

Defining closure as any grid cell with 25% overlap of the polygon. This step also establishes the surrounding stat_areas to use for the analysis, based on a buffer parameter chosen (here, 200 km).

```{r}

#| label: closure-setup

# 1. Load the polygons using terra
closure_polygon       <- terra::vect(paste0("shapefiles/", params$closure, ".shp"))
stat_areas_with_area  <- stat_areas

# Ensure character type right away and calculate individual cell base areas
stat_areas_with_area$STAT_AREA <- trimws(as.character(stat_areas_with_area$STAT_AREA))
stat_areas_with_area$cell_base_area <- terra::expanse(stat_areas_with_area)

# 2. Project stat areas if they don't match the closure polygon CRS
if (crs(stat_areas_with_area) != crs(closure_polygon)) {
  stat_areas_with_area <- terra::project(stat_areas_with_area, crs(closure_polygon))
}

# 3. Build a geometric area overlap intersection
intersected_geometry <- terra::intersect(stat_areas_with_area, closure_polygon)

# 4. FIXED: Calculate what % of the LOCAL STAT AREA CELL is covered by the closure
intersected_geometry$overlap_area <- terra::expanse(intersected_geometry)
intersected_geometry$pct_of_cell_covered <- (intersected_geometry$overlap_area / intersected_geometry$cell_base_area) * 100

# 5. Extract valid stat areas based on local coverage 
# (Using 1.0% here catches any cell where the closure cuts through a meaningful chunk)
threshold_pct       <- 25.0
valid_intersections <- as.data.frame(intersected_geometry) %>% 
  filter(pct_of_cell_covered > threshold_pct)

valid_stat_areas   <- unique(as.character(valid_intersections$STAT_AREA))
closure.stat.areas <- trimws(valid_stat_areas)

# 6. Separate the dataset into closure and non-closure polygons
is_overlapping <- stat_areas_with_area$STAT_AREA %in% valid_stat_areas

# Isolate statistical areas meeting the threshold requirements
closure        <- stat_areas_with_area[is_overlapping, ]
closure$closed <- "yes"
closure2       <- terra::aggregate(closure)

# Isolate statistical areas that fell below the threshold requirements
non_closure        <- stat_areas_with_area[!is_overlapping, ]
non_closure$closed <- "no"

# ==============================================================================
# Generate the 100 km spatial buffer around the true aggregated closure
# ==============================================================================
closure_buffer <- terra::buffer(closure2, width = params$buffer) 

# Identify which stat areas fall within this 100 km window
local_stat_matches <- terra::relate(stat_areas_with_area, closure_buffer, "intersects")
local_stat_areas   <- stat_areas_with_area$STAT_AREA[as.vector(local_stat_matches)]

# 7. Quick visual validation check

# Subset local stat areas for plotting
local_stats_spat <- stat_areas_with_area[stat_areas_with_area$STAT_AREA %in% local_stat_areas, ]

ggplot() +
  # 1. Base Stat Areas layer
  tidyterra::geom_spatvector(
    data = local_stats_spat, 
    fill = NA, 
    color = "gray80", 
    lwd = 0.3
  ) +
  # 2. Closure layer (Blue)
  tidyterra::geom_spatvector(
    data = closure, 
    aes(color = "Closure (Stat Area)"), 
    fill = NA, 
    lwd = 1.0
  ) +
  # 3. Closure Polygon layer (Red)
  tidyterra::geom_spatvector(
    data = closure_polygon, 
    aes(color = "Closure Polygon"), 
    fill = NA, 
    lwd = 0.8
  ) +
  # 4. Closure Buffer layer (Purple Dashed)
  tidyterra::geom_spatvector(
    data = closure_buffer, 
    aes(color = "Closure Buffer"), 
    fill = NA, 
    lwd = 0.8, 
    linetype = "dotted"
  ) +
  # Manual Scale for Clean Legend Mapping
  scale_color_manual(
    name = "Spatial Layer",
    values = c(
      "Closure (Stat Area)"         = "blue",
      "Closure Polygon" = "red",
      "Closure Buffer"  = "purple"
    )
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "right",                       # Cleanly places legend on the right canvas margin
    legend.background = element_rect(fill = "white", color = "gray80"),
    axis.title = element_blank()
  )

```

## 5. Catch Data

Load catch data from the catch accounting system (CAS) and assign to stat_areas.

```{r}
# ==============================================================================
# SECTION 5: CATCH DATA LOAD & DOUBLE-SIDED 1%/99% OUTLIER CAPPING
# ==============================================================================
#| label: catch-data

suppressPackageStartupMessages({
  library(dplyr)
  library(lubridate)
  library(stringr)
})

# 1. Read standalone species file
catch_data_raw <- read.csv(paste0("data/catch_data_", params$psc_spp, ".csv"), header = TRUE)
colnames(catch_data_raw) <- toupper(colnames(catch_data_raw))

# Capture time step string for clean conditional evaluation
time_step_mode <- params$time_step

# 2. Slice out matching parameters and assign standardized TIME_STEP_KEY
catch_data_filtered <- catch_data_raw %>%
  dplyr::filter(
    YEAR %in% params$timeframe,
    SPECIES_GROUP_CODE == toupper(params$psc_spp),
    AGENCY_GEAR_CODE == params$gear,
    HARVEST_SECTOR == params$sector  # Dynamic Sector Filter
  ) %>%
  dplyr::mutate(
    STAT_AREA = stringr::str_pad(trimws(as.character(STAT_AREA)), width = 6, side = "left", pad = "0"),
    
    CATCH_ACTIVITY_DATE = as.Date(CATCH_ACTIVITY_DATE),
    WEEK_END_DATE       = as.Date(WEEK_END_DATE),
    
    # Modern conditional syntax (No deprecation warnings)
    TIME_STEP_KEY = if (time_step_mode == "annual") {
      "Full-Year"
    } else if (time_step_mode == "monthly") {
      sprintf("M%02d", lubridate::month(CATCH_ACTIVITY_DATE))
    } else {
      sprintf("W%02d", lubridate::isoweek(CATCH_ACTIVITY_DATE))
    }
  )

# 3. CALCULATE 1ST AND 99TH PERCENTILE THRESHOLDS
gf_p01 <- quantile(catch_data_filtered$GF_TOTAL_CATCH_MT, 0.01, na.rm = TRUE)
gf_p99 <- quantile(catch_data_filtered$GF_TOTAL_CATCH_MT, 0.99, na.rm = TRUE)

psc_p01 <- quantile(catch_data_filtered$PSC_VALUE, 0.01, na.rm = TRUE)
psc_p99 <- quantile(catch_data_filtered$PSC_VALUE, 0.99, na.rm = TRUE)

# 4. CAP UPPER & LOWER 99% PERCENTILES (WINSORIZE GF AND PSC)
catch_data <- catch_data_filtered %>%
  dplyr::mutate(
    # Cap GF Catch
    GF_TOTAL_CATCH_MT = pmax(gf_p01, pmin(GF_TOTAL_CATCH_MT, gf_p99)),
    
    # Cap PSC Catch
    PSC_VALUE         = pmax(psc_p01, pmin(PSC_VALUE, psc_p99))
  )

# 5. Collapse daily entries while preserving NA_real_ for unobserved effort
catch_data_binned <- catch_data %>%
  dplyr::group_by(YEAR, TIME_STEP_KEY, STAT_AREA, HARVEST_SECTOR, SPECIES_GROUP_CODE) %>%
  dplyr::summarise(
    GF_TOTAL_CATCH_MT   = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE),
    PSC_VALUE           = sum(PSC_VALUE, na.rm = TRUE),
    EFFORT_MINUTES      = ifelse(all(is.na(EFFORT_MINUTES)), NA_real_, sum(EFFORT_MINUTES, na.rm = TRUE)),
    CATCH_ACTIVITY_DATE = min(CATCH_ACTIVITY_DATE),
    .groups = "drop"
  )

# 6. Align spatial object and merge
stat_areas_with_area$STAT_AREA <- stringr::str_pad(trimws(as.character(stat_areas_with_area$STAT_AREA)), width = 6, side = "left", pad = "0")

catch_data_full <- merge(stat_areas_with_area, catch_data_binned, by = "STAT_AREA", all.x = FALSE)
catch_data2     <- subset(catch_data_full, catch_data_full$STAT_AREA %in% local_stat_areas)

```

## 6. Inspect effort and catch

Groundfish catch.
```{r}

#| label: fig-gf-catch
#| fig-cap: "Groundfish catch distributions."
#| fig-width: 12
#| fig-height: 8.5

ggplot() +
  tidyterra::geom_spatvector(data = subset(catch_data2, catch_data2$YEAR %in% 2014), aes(fill = GF_TOTAL_CATCH_MT), 
                             alpha = 0.8, col = NA, na.rm = TRUE) +
  scale_fill_distiller(palette = "GnBu", name = "GF (mt)") +
  tidyterra::geom_spatvector(data = closure2, aes(color = "Closure"), 
                             fill = NA, alpha = 0.5, lwd = 0.5) +
  scale_color_manual(values = c("Closure" = "red"), name = NULL) +
facet_wrap(
    vars(TIME_STEP_KEY), 
    nrow = 5,
    ncol=10
  )
  theme_bw() + theme(plot.title = element_text(hjust = 0.5), axis.title = element_blank()) +
  theme(
    axis.title = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank()
  )

```

PSC catch.
```{r}

#| label: fig-psc-catch
#| fig-cap: "PSC catch distributions."
#| fig-width: 12
#| fig-height: 8.5

ggplot() +
  tidyterra::geom_spatvector(data = subset(catch_data2, catch_data2$YEAR %in% 2014), aes(fill = PSC_VALUE), 
                             alpha = 0.8, col = NA, na.rm = TRUE) +
  scale_fill_distiller(palette = "GnBu", name = paste0(spp_title, " PSC (num)")) +
  tidyterra::geom_spatvector(data = closure2, aes(color = "Closure"), 
                             fill = NA, alpha = 0.5, lwd = 0.5) +
  scale_color_manual(values = c("Closure" = "red"), name = NULL) +
facet_wrap(
    vars(TIME_STEP_KEY), 
    nrow = 5,
    ncol=10
  )
  theme_bw() + theme(plot.title = element_text(hjust = 0.5), axis.title = element_blank()) +
  theme(
    axis.title = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank()
  )

```
Effort.
```{r}

#| label: fig-gf-effort
#| fig-cap: "Groundfish Effort distributions."
#| fig-width: 12
#| fig-height: 8.5

ggplot() +
  tidyterra::geom_spatvector(data = subset(catch_data2, catch_data2$YEAR %in% 2014), aes(fill = EFFORT_MINUTES), 
                             alpha = 0.8, col = NA, na.rm = TRUE) +
  scale_fill_distiller(palette = "GnBu", name = "Effort (min") +
  tidyterra::geom_spatvector(data = closure2, aes(color = "Closure"), 
                             fill = NA, alpha = 0.5, lwd = 0.5) +
  scale_color_manual(values = c("Closure" = "red"), name = NULL) +
facet_wrap(
    vars(TIME_STEP_KEY), 
    nrow = 5,
    ncol=10
  )
  theme_bw() + theme(plot.title = element_text(hjust = 0.5), axis.title = element_blank()) +
  theme(
    axis.title = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank()
  )

```
## 7. Inspect the temporal autocorrelation assumption

Sufficient temporal autocorrelation from one year to the next. 
```{r}

#| label: fig-global-ar1
#| fig-cap: "Global temporal autocorrelation in groundfish catch."
#| fig-width: 12
#| fig-height: 8.5

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(scales)
  library(patchwork)
})

# ==============================================================================
# 1. COMPUTE DERIVED PROPORTIONS AND RATES
# ==============================================================================
df_primitives_clean <- catch_data_binned %>%
  dplyr::filter(
    YEAR != 2020,
    STAT_AREA %in% local_stat_areas
  ) %>%
  as.data.frame()

# Calculate cell-level rates and regional proportions
df_derived_rates <- df_primitives_clean %>%
  # Sum regional totals per year and sector for proportion calculation
  dplyr::group_by(YEAR, HARVEST_SECTOR) %>%
  dplyr::mutate(regional_gf_total = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE)) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(YEAR, STAT_AREA, HARVEST_SECTOR) %>%
  dplyr::summarise(
    # 1. Spatial Redistribution Share (Cell MT / Regional Total MT)
    gf_prop = ifelse(first(regional_gf_total) > 0, 
                     sum(GF_TOTAL_CATCH_MT, na.rm = TRUE) / first(regional_gf_total), 0),
    
    # 2. Method 1 Bycatch Rate (PSC Count / Groundfish MT)
    psc_rate_mt = ifelse(sum(GF_TOTAL_CATCH_MT, na.rm = TRUE) > 0, 
                         sum(PSC_VALUE, na.rm = TRUE) / sum(GF_TOTAL_CATCH_MT, na.rm = TRUE), 0),
    
    # 3. Method 2 CPUE (Groundfish MT / Effort Minutes)
    cpue_mt_min = ifelse(sum(EFFORT_MINUTES, na.rm = TRUE) > 0, 
                         sum(GF_TOTAL_CATCH_MT, na.rm = TRUE) / sum(EFFORT_MINUTES, na.rm = TRUE), 0),
    .groups = "drop"
  )

# ==============================================================================
# 2. MATCH LAGGED (t-1) VS ACTUAL (t) PARAMETERS
# ==============================================================================
rates_lagged <- df_derived_rates %>%
  dplyr::rename(
    prop_prior = gf_prop,
    rate_prior = psc_rate_mt,
    cpue_prior = cpue_mt_min
  ) %>%
  dplyr::mutate(VALIDATION_YEAR_MATCH = YEAR + 1)

rates_actual <- df_derived_rates %>%
  dplyr::rename(
    prop_actual = gf_prop,
    rate_actual = psc_rate_mt,
    cpue_actual = cpue_mt_min
  )

global_rates_panel <- rates_actual %>%
  dplyr::inner_join(
    rates_lagged,
    by = c("YEAR" = "VALIDATION_YEAR_MATCH", "STAT_AREA", "HARVEST_SECTOR")
  )

# ==============================================================================
# 3. HELPER PLOTTING FUNCTION FOR SINGLE GLOBAL SCATTER
# ==============================================================================
make_rate_scatter_global <- function(df, x_var, y_var, title_text, x_lab, y_lab) {
  
  # Calculate overall global correlations across all years
  r_val   <- suppressWarnings(cor(df[[x_var]], df[[y_var]], use = "complete.obs", method = "pearson"))
  rho_val <- suppressWarnings(cor(df[[x_var]], df[[y_var]], use = "complete.obs", method = "spearman"))
  
  label_text <- sprintf("Global r = %.2f\nGlobal rho = %.2f", 
                        ifelse(is.na(r_val), 0, r_val), 
                        ifelse(is.na(rho_val), 0, rho_val))
  
  # Set text position dynamically at 90% Y max and min X
  x_pos <- min(df[[x_var]], na.rm = TRUE)
  y_pos <- max(df[[y_var]], na.rm = TRUE) * 0.90

  ggplot(df, aes(x = .data[[x_var]], y = .data[[y_var]])) +
    geom_point(aes(color = HARVEST_SECTOR), size = 2, alpha = 0.6) +
    geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "solid", linewidth = 0.8) +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
    annotate(
      "text",
      x = x_pos, y = y_pos, label = label_text,
      hjust = 0, fontface = "bold", size = 3.5, color = "navy"
    ) +
    scale_x_continuous(labels = scales::comma) +
    scale_y_continuous(labels = scales::comma) +
    scale_color_manual(values = c("CP" = "#377eb8", "CV" = "#e41a1c")) +
    theme_bw() +
    labs(
      title = title_text,
      x = x_lab,
      y = y_lab,
      color = "Sector"
    ) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5, size = 11),
      axis.title = element_text(size = 9.5),
      panel.grid.minor = element_blank()
    )
}

# ==============================================================================
# 4. GENERATE THE GLOBAL SCATTER PLOT
# ==============================================================================
p1 <- make_rate_scatter_global(
  global_rates_panel, "prop_prior", "prop_actual",
  "GF Spatial Redistribution Share", "Prior Share (t-1)", "Actual Share (t)"
)

# Display single plot with title annotation
combined_rates_panel <- p1 +
  plot_annotation(
    title = paste0("Derived Parameter Autocorrelation Validation: ", title_name),
    subtitle = "Cell-Level Rates & Spatial Allocations: Year t-1 vs Year t (All Years Combined)",
    theme = theme(
      plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
      plot.subtitle = element_text(face = "italic", size = 10, hjust = 0.5)
    )
  ) & 
  theme(legend.position = "bottom")

# Display panel
print(combined_rates_panel)

```

By year, sufficient high temporal autocorrelation.
```{r}

#| label: fig-global-ar1
#| fig-cap: "Annual temporal autocorrelation in groundfish spatial catch distribution."
#| fig-width: 10
#| fig-height: 10

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(scales)
})

# ==============================================================================
# 1. COMPUTE DERIVED GROUNDFISH SPATIAL PROPORTIONS
# ==============================================================================
df_primitives_clean <- catch_data_binned %>%
  dplyr::filter(
    YEAR != 2020,
    STAT_AREA %in% local_stat_areas,
    HARVEST_SECTOR == params$sector # Dynamic sector binding
  ) %>%
  as.data.frame()

# Calculate cell-level groundfish redistribution proportions
df_derived_rates <- df_primitives_clean %>%
  dplyr::group_by(YEAR, HARVEST_SECTOR) %>%
  dplyr::mutate(regional_gf_total = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE)) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(YEAR, STAT_AREA, HARVEST_SECTOR) %>%
  dplyr::summarise(
    # Spatial Redistribution Share (Cell MT / Regional Total MT)
    gf_prop = ifelse(first(regional_gf_total) > 0, 
                     sum(GF_TOTAL_CATCH_MT, na.rm = TRUE) / first(regional_gf_total), 0),
    .groups = "drop"
  )

# ==============================================================================
# 2. MATCH LAGGED (t-1) VS ACTUAL (t) PROPORTIONS
# ==============================================================================
rates_lagged <- df_derived_rates %>%
  dplyr::rename(prop_prior = gf_prop) %>%
  dplyr::mutate(VALIDATION_YEAR_MATCH = YEAR + 1)

rates_actual <- df_derived_rates %>%
  dplyr::rename(prop_actual = gf_prop)

global_rates_panel <- rates_actual %>%
  dplyr::inner_join(
    rates_lagged,
    by = c("YEAR" = "VALIDATION_YEAR_MATCH", "STAT_AREA", "HARVEST_SECTOR")
  )

# ==============================================================================
# 3. COMPUTE YEAR-BY-YEAR CORRELATIONS
# ==============================================================================
annual_corrs_rates <- global_rates_panel %>%
  dplyr::group_by(YEAR) %>%
  dplyr::summarise(
    prop_r   = suppressWarnings(cor(prop_prior, prop_actual, use = "complete.obs", method = "pearson")),
    prop_rho = suppressWarnings(cor(prop_prior, prop_actual, use = "complete.obs", method = "spearman")),
    .groups = "drop"
  )

global_rates_panel <- global_rates_panel %>%
  dplyr::left_join(annual_corrs_rates, by = "YEAR")

# ==============================================================================
# 4. PLOT CLEAN FACETED SCATTER (SQUARE PANELS & ANCHORED LABELS)
# ==============================================================================
max_y_limit <- max(global_rates_panel$prop_actual, na.rm = TRUE) * 1.05

labels_df <- global_rates_panel %>%
  dplyr::group_by(YEAR) %>%
  dplyr::summarise(
    r_val   = first(prop_r),
    rho_val = first(prop_rho),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    label_text = sprintf("r = %.2f\n\u03c1 = %.2f", 
                         ifelse(is.na(r_val), 0, r_val), 
                         ifelse(is.na(rho_val), 0, rho_val))
  )

ggplot(global_rates_panel, aes(x = prop_prior, y = prop_actual)) +
  geom_point(aes(color = HARVEST_SECTOR), size = 2, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "solid", linewidth = 0.8) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  
  # ANCHOR TEXT AT TOP-LEFT ABOVE DATA CLUSTER
  geom_text(
    data = labels_df,
    aes(x = 0.01, y = max_y_limit * 0.88, label = label_text),
    hjust = 0, vjust = 1, fontface = "bold", size = 3.2, color = "navy"
  ) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, max_y_limit)) +
  scale_color_manual(values = c("CP" = "#377eb8", "CV" = "#e41a1c")) +
  theme_bw() +
  labs(
    title = paste0("Annual Autocorrelation Validation: ", title_name),
    subtitle = "Groundfish Spatial Redistribution Share: Year t-1 vs Year t",
    x = "Prior Share (t-1)",
    y = "Actual Share (t)",
    color = "Sector"
  ) +
  theme(
    plot.title       = element_text(face = "bold", hjust = 0.5, size = 12),
    plot.subtitle    = element_text(face = "italic", hjust = 0.5, size = 10),
    axis.title       = element_text(size = 9.5),
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    
    # FORCES SQUARE FACET PANELS & SPACING
    aspect.ratio     = 1,
    panel.spacing    = unit(0.6, "lines")
  ) + 
  facet_wrap(~YEAR, ncol = 5)

```
## 8. Assign Closure Dates & Isolate Impacted Hauls

```{r}

#| label: impacted-hauls
#| echo: false

# 1. Sum status quo PSC by year for baseline reality joins
all_psc_status_quo_annual <- as.data.frame(catch_data2) %>%
  dplyr::group_by(YEAR) %>%
  dplyr::reframe(
    sum_psc_status_quo = sum(PSC_VALUE, na.rm = TRUE)
  )

# 2. Assign closure dates matching your global parameter timeline
closure_dates <- data.frame(
  YEAR = params$timeframe,
  closure_date = as.Date(paste0(params$timeframe, "-01-01"))
)

all_psc_closures_dates <- as.data.frame(dplyr::left_join(catch_data2, closure_dates, by = "YEAR"))

# Flag closure presence and catch activity dates
all_psc_closures_dates$closure  <- ifelse(!is.na(all_psc_closures_dates$closure_date), "yes", "no")
all_psc_closures_dates$impacted <- ifelse(all_psc_closures_dates$CATCH_ACTIVITY_DATE >= all_psc_closures_dates$closure_date, "yes", "no")

all_psc_closures_dates$tons_gf_after_closure <- ifelse(all_psc_closures_dates$impacted == "yes", all_psc_closures_dates$GF_TOTAL_CATCH_MT, 0)
all_psc_closures_dates$tons_gf_no_closure    <- ifelse(all_psc_closures_dates$impacted == "no" | is.na(all_psc_closures_dates$impacted), all_psc_closures_dates$GF_TOTAL_CATCH_MT, 0)

# Isolate impacted hauls
all_psc_impacted <- subset(all_psc_closures_dates, impacted == "yes")

# 3. Clean Ledger Prep for validation blocks
ledger_df_clean <- as.data.frame(all_psc_closures_dates) %>%
  dplyr::mutate(
    STAT_AREA = stringr::str_pad(trimws(as.character(STAT_AREA)), width = 6, side = "left", pad = "0"),
    inside    = ifelse(STAT_AREA %in% closure.stat.areas, "yes", "no")
  ) %>%
  dplyr::filter(STAT_AREA %in% local_stat_areas, inside == "no")

# 4. Prepare total.join.plot.annual2 as the baseline reality bridge
total.join.plot.annual2 <- all_psc_status_quo_annual %>%
  dplyr::mutate(
    psc_statusquo   = sum_psc_status_quo,
    psc_change_m1   = 0,
    psc_change_m2   = 0,
    net_new_PSC_m1  = sum_psc_status_quo,
    net_new_PSC_m2  = sum_psc_status_quo,
    LABEL           = "SPECIES"
  ) %>%
  dplyr::select(YEAR, psc_statusquo, psc_change_m1, psc_change_m2, net_new_PSC_m1, net_new_PSC_m2, LABEL)

```

# RETROSPECTIVE "WHAT-IF" SCENARIO
This scenario runs the condition of "what if a closure had occurred" at the given year and time step given. 

## 9. Calculate Catch and Proportions Inside/Outside Closure

This step summarizes the data used in the main calculations - the proportions of outside groundish catch where displaced catch is applied to, under the assumption that the fishery will meet the same overall total catch, the total catch inside the closure to be applied to the proportions, the new PSC as a result of the new GF catch applied to local PSC rates, and the PSC caught inside the closure that will be subtracted from the net total.

```{r}

#| label: rates-props-5yr-cpue
#| echo: false

stat_areas_closed <- as.character(closure.stat.areas) 

all_psc_impacted_cluster <- all_psc_impacted %>%
  mutate(inside = case_when(trimws(as.character(STAT_AREA)) %in% stat_areas_closed ~ "yes", TRUE ~ "no"))

# --- STEP A: INSIDE CATCH SUMMARY ---
gf_inside_cluster2 <- all_psc_impacted_cluster %>%
  dplyr::filter(inside == "yes") %>%
  dplyr::group_by(YEAR, TIME_STEP_KEY, HARVEST_SECTOR) %>%
  dplyr::summarise(
    tons_gf_inside_wk     = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE),
    psc_measure_inside_wk = sum(PSC_VALUE, na.rm = TRUE),
    effort_inside_wk      = sum(EFFORT_MINUTES, na.rm = TRUE), 
    .groups = "drop"
  )

# --- STEP B: OUTSIDE PERFORMANCE MATRIX ---
gf_outside_cluster <- all_psc_impacted_cluster %>%
  filter(inside == "no") %>%
  mutate(DYNAMIC_SEASONAL_KEY = TIME_STEP_KEY) # Standardized key reuse!

# 1. 5-YEAR ROLLING MEAN CPUE
annual_spatial_footprint <- gf_outside_cluster %>%
  group_by(YEAR, DYNAMIC_SEASONAL_KEY, STAT_AREA, HARVEST_SECTOR) %>%
  summarise(
    outside_gf_tons     = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE),
    outside_effort_mins = sum(EFFORT_MINUTES, na.rm = TRUE),
    .groups = "drop"
  )

rolling_cpue_list <- list()
all_years_present <- sort(unique(annual_spatial_footprint$YEAR))

for(target_yr in all_years_present) {
  window_start <- target_yr - 1
  window_end   <- target_yr - 1
  
  rolling_slice <- annual_spatial_footprint %>%
    dplyr::filter(YEAR >= window_start & YEAR <= window_end) %>%
    dplyr::group_by(DYNAMIC_SEASONAL_KEY, STAT_AREA, HARVEST_SECTOR) %>%
    dplyr::summarise(
      sum_tons_5yr   = sum(outside_gf_tons, na.rm = TRUE),
      sum_effort_5yr = sum(outside_effort_mins, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      hist_gf_cpue_seasonal = ifelse(sum_effort_5yr > 0, sum_tons_5yr / sum_effort_5yr, 0),
      YEAR = target_yr
    ) %>%
    dplyr::select(YEAR, DYNAMIC_SEASONAL_KEY, STAT_AREA, HARVEST_SECTOR, hist_gf_cpue_seasonal)
  
  rolling_cpue_list[[as.character(target_yr)]] <- rolling_slice
}

historical_cpue_profile <- do.call(rbind, rolling_cpue_list)

# 2. ACTIVE BASELINE MATRIX
weekly_base_metrics <- gf_outside_cluster %>%
  group_by(YEAR, TIME_STEP_KEY, DYNAMIC_SEASONAL_KEY, STAT_AREA, HARVEST_SECTOR) %>%
  summarise(
    tons_gf_wk           = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE),
    total_effort_minutes = sum(EFFORT_MINUTES, na.rm = TRUE),
    total_psc_count      = sum(PSC_VALUE, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    mean_psc_rate_wk         = ifelse(tons_gf_wk > 0, total_psc_count / tons_gf_wk, 0),
    mean_psc_rate_minutes_wk = ifelse(total_effort_minutes > 0, total_psc_count / total_effort_minutes, 0),
    active_gf_cpue_wk        = ifelse(total_effort_minutes > 0, tons_gf_wk / total_effort_minutes, 0)
  ) %>%
  left_join(
    gf_outside_cluster %>%
      group_by(YEAR, TIME_STEP_KEY, HARVEST_SECTOR) %>%
      summarise(
        tons_gf_annual      = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE),
        effort_annual_total = sum(EFFORT_MINUTES, na.rm = TRUE),
        .groups = "drop"
      ),
    by = c("YEAR", "TIME_STEP_KEY", "HARVEST_SECTOR")
  ) %>%
  left_join(
    gf_outside_cluster %>%
      group_by(YEAR, TIME_STEP_KEY, STAT_AREA, HARVEST_SECTOR) %>%
      summarise(
        tons_gf_annual_stat   = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE),
        effort_annual_stat    = sum(EFFORT_MINUTES, na.rm = TRUE),
        .groups = "drop"
      ),
    by = c("YEAR", "TIME_STEP_KEY", "STAT_AREA", "HARVEST_SECTOR")
  ) %>%
  mutate(
    proportion_outside_wk        = ifelse(tons_gf_annual > 0, tons_gf_annual_stat / tons_gf_annual, 0),
    proportion_outside_effort_wk = ifelse(effort_annual_total > 0, effort_annual_stat / effort_annual_total, 0),
    area_cpue                     = ifelse(effort_annual_stat > 0, tons_gf_annual_stat / effort_annual_stat, 0)
  ) %>%
  left_join(
    gf_outside_cluster %>%
      group_by(YEAR, TIME_STEP_KEY, STAT_AREA, HARVEST_SECTOR) %>%
      summarise(
        stat_tons = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE),
        stat_eff  = sum(EFFORT_MINUTES, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      mutate(stat_cpue = ifelse(stat_eff > 0, stat_tons / stat_eff, 0)) %>%
      group_by(YEAR, TIME_STEP_KEY, HARVEST_SECTOR) %>%
      summarise(total_outside_cpue_sum = sum(stat_cpue, na.rm = TRUE), .groups = "drop"),
    by = c("YEAR", "TIME_STEP_KEY", "HARVEST_SECTOR")
  ) %>%
  mutate(
    proportion_outside_cpue_wk = ifelse(total_outside_cpue_sum > 0, area_cpue / total_outside_cpue_sum, 0)
  ) %>%
  select(-area_cpue, -total_outside_cpue_sum)

# 3. JOIN 5-YEAR CPUE
sum.stat.timeseries.gf_all3 <- weekly_base_metrics %>%
  left_join(
    historical_cpue_profile,
    by = c("YEAR", "DYNAMIC_SEASONAL_KEY", "STAT_AREA", "HARVEST_SECTOR")
  ) %>%
  mutate(
    hist_gf_cpue_seasonal = ifelse(!is.na(hist_gf_cpue_seasonal), hist_gf_cpue_seasonal, active_gf_cpue_wk)
  ) %>%
  as.data.frame()

```


## 10. Calculate Weekly and Annual Summary Statistics

This step applies the displacement and new PSC calculations.

```{r}

# --- SECTION 10: CORE PIPELINE ---
total.join.master <- sum.stat.timeseries.gf_all3 %>%
  left_join(
    gf_inside_cluster2 %>% select(YEAR, TIME_STEP_KEY, HARVEST_SECTOR, tons_gf_inside_wk),
    by = c("YEAR", "TIME_STEP_KEY", "HARVEST_SECTOR")
  ) %>%
  mutate(
    tons_gf_inside_wk = replace_na(tons_gf_inside_wk, 0),
    
    # Shared Retrospective Catch Allocation
    gf_catch_new = tons_gf_inside_wk * proportion_outside_wk,
    
    # Method 1: Volume Track (Real-Time Catch * Real-Time Bycatch-Per-Ton)
    new_PSC_method1 = gf_catch_new * mean_psc_rate_wk,
    
    # Method 2: Effort Track (Shared Tonnage / 1-Yr Lagged CPUE * Real-Time Bycatch-Per-Minute)
    effort_minutes_needed = ifelse(hist_gf_cpue_seasonal > 0, gf_catch_new / hist_gf_cpue_seasonal, 0),
    new_PSC_method2       = effort_minutes_needed * mean_psc_rate_minutes_wk
  )

# Roll up to the annual summary as standard
ledger_weekly <- total.join.master %>%
  group_by(YEAR, TIME_STEP_KEY, HARVEST_SECTOR) %>%
  summarise(
    new_PSC_m1_wk = sum(new_PSC_method1, na.rm = TRUE),
    new_PSC_m2_wk = sum(new_PSC_method2, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(
    gf_inside_cluster2 %>% select(YEAR, TIME_STEP_KEY, HARVEST_SECTOR, psc_measure_inside_wk),
    by = c("YEAR", "TIME_STEP_KEY", "HARVEST_SECTOR")
  ) %>%
  mutate(
    psc_measure_inside_wk = replace_na(psc_measure_inside_wk, 0),
    net_change_m1_wk = new_PSC_m1_wk - psc_measure_inside_wk,
    net_change_m2_wk = new_PSC_m2_wk - psc_measure_inside_wk
  )

total.join.plot.annual <- ledger_weekly %>%
  group_by(YEAR) %>%
  summarise(
    est_change_method1 = sum(net_change_m1_wk, na.rm = TRUE),
    est_change_method2 = sum(net_change_m2_wk, na.rm = TRUE),
    .groups = "drop"
  )

total.join = total.join.master

```

## 11. Bootstrap

To provide a measure of uncertainty, we are bootstrapping 1,000 alternative scenarios by randomly shuffling the PSC rates where fishing is redistributed. We calculated a 95% Confidence Interval using the middle 95% of all simulated outcomes (between 2.5th and 97.5th percentiles).

```{r}

# ==============================================================================
# ## 11. Bootstrap (ROW-PRESERVING COUPLED RESAMPLER)
# ==============================================================================
#| label: spatial-rate-bootstrap

set.seed(345)
n_iterations <- 1000

df_bs_base <- total.join %>%
  dplyr::left_join(
    gf_inside_cluster2 %>% dplyr::select(YEAR, TIME_STEP_KEY, HARVEST_SECTOR, psc_measure_inside_wk),
    by = c("YEAR", "TIME_STEP_KEY", "HARVEST_SECTOR")
  ) %>%
  dplyr::mutate(
    psc_measure_inside_wk = tidyr::replace_na(psc_measure_inside_wk, 0),
    STRATA_KEY = paste0(YEAR, "_", TIME_STEP_KEY, "_", HARVEST_SECTOR) # Include YEAR in strata key!
  )

strata_steps <- unique(df_bs_base$STRATA_KEY)

boot_matrix_m1 <- matrix(NA, nrow = n_iterations, ncol = length(strata_steps))
boot_matrix_m2 <- matrix(NA, nrow = n_iterations, ncol = length(strata_steps))
colnames(boot_matrix_m1) <- strata_steps
colnames(boot_matrix_m2) <- strata_steps

for(i in 1:n_iterations) {
  new_psc_m1_iteration <- numeric(nrow(df_bs_base))
  new_psc_m2_iteration <- numeric(nrow(df_bs_base))
  
  for(st in strata_steps) {
    st_indices <- which(df_bs_base$STRATA_KEY == st)
    n_records  <- length(st_indices)
    
    if(n_records > 0) {
      sampled_rows <- sample(st_indices, size = n_records, replace = TRUE)
      
      # Method 1 Bootstrap Evaluation
      new_psc_m1_iteration[st_indices] <- df_bs_base$gf_catch_new[st_indices] * df_bs_base$mean_psc_rate_wk[sampled_rows]
      
      # Method 2 Bootstrap Evaluation
      gf_shared_tons <- df_bs_base$gf_catch_new[st_indices]
      cpue_m2        <- df_bs_base$hist_gf_cpue_seasonal[sampled_rows]
      rates_m2       <- df_bs_base$mean_psc_rate_minutes_wk[sampled_rows]
      
      mins_needed <- ifelse(cpue_m2 > 0, gf_shared_tons / cpue_m2, 0)
      new_psc_m2_iteration[st_indices] <- mins_needed * tidyr::replace_na(rates_m2, 0)
    }
  }
  
  iter_summary <- df_bs_base %>%
    dplyr::mutate(boot_m1 = new_psc_m1_iteration, boot_m2 = new_psc_m2_iteration) %>%
    dplyr::group_by(YEAR, TIME_STEP_KEY, HARVEST_SECTOR, STRATA_KEY) %>%
    dplyr::summarise(
      sim_outside_m1 = sum(boot_m1, na.rm = TRUE),
      sim_outside_m2 = sum(boot_m2, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    # INCLUDE YEAR IN THIS JOIN TO AVOID MULTI-YEAR MANY-TO-MANY DUPLICATION
    dplyr::left_join(
      gf_inside_cluster2 %>% dplyr::select(YEAR, TIME_STEP_KEY, HARVEST_SECTOR, psc_measure_inside_wk),
      by = c("YEAR", "TIME_STEP_KEY", "HARVEST_SECTOR")
    ) %>%
    dplyr::mutate(
      psc_measure_inside_wk = tidyr::replace_na(psc_measure_inside_wk, 0),
      net_change_m1 = sim_outside_m1 - psc_measure_inside_wk,
      net_change_m2 = sim_outside_m2 - psc_measure_inside_wk
    )
  
  for(j in seq_along(strata_steps)) {
    match_idx <- which(iter_summary$STRATA_KEY == strata_steps[j])
    if(length(match_idx) == 1) { # Strictly 1 match per unique stratum
      boot_matrix_m1[i, j] <- iter_summary$net_change_m1[match_idx]
      boot_matrix_m2[i, j] <- iter_summary$net_change_m2[match_idx]
    }
  }
}

# Long transformations
boot_long_m1 <- as.data.frame(boot_matrix_m1) %>% dplyr::mutate(iteration = 1:n()) %>%
  tidyr::pivot_longer(-iteration, names_to = "STRATA_KEY", values_to = "net_change")

boot_long_m2 <- as.data.frame(boot_matrix_m2) %>% dplyr::mutate(iteration = 1:n()) %>%
  tidyr::pivot_longer(-iteration, names_to = "STRATA_KEY", values_to = "net_change")

strata_year_map <- df_bs_base %>% dplyr::select(STRATA_KEY, YEAR) %>% dplyr::distinct()

boot_annual_m1 <- boot_long_m1 %>% dplyr::left_join(strata_year_map, by = "STRATA_KEY") %>%
  dplyr::group_by(iteration, YEAR) %>% dplyr::summarise(annual_change = sum(net_change, na.rm = TRUE), .groups = "drop")

boot_annual_m2 <- boot_long_m2 %>% dplyr::left_join(strata_year_map, by = "STRATA_KEY") %>%
  dplyr::group_by(iteration, YEAR) %>% dplyr::summarise(annual_change = sum(net_change, na.rm = TRUE), .groups = "drop")

boot_m1_ci <- boot_annual_m1 %>% dplyr::group_by(YEAR) %>%
  dplyr::summarise(lower_change_m1 = quantile(annual_change, 0.025, na.rm = TRUE),
                   upper_change_m1 = quantile(annual_change, 0.975, na.rm = TRUE), .groups = "drop")

boot_m2_ci <- boot_annual_m2 %>% dplyr::group_by(YEAR) %>%
  dplyr::summarise(lower_change_m2 = quantile(annual_change, 0.025, na.rm = TRUE),
                   upper_change_m2 = quantile(annual_change, 0.975, na.rm = TRUE), .groups = "drop")

boot_annual_ci <- dplyr::left_join(boot_m1_ci, boot_m2_ci, by = "YEAR")

```


## 12. Evaluate Impacts

Compares the estimated PSC change to the observed PSC (status quo) each year.

```{r}

# ==============================================================================
# ## 12. Evaluate Impacts
# ==============================================================================
#| label: fig-status-quo
#| fig-width: 12
#| fig-height: 6

# 1. Clean parameter labels to avoid any trailing workspace text mismatches
clean_spp <- trimws(toupper(params$psc_spp))
spp_title_fixed <- case_when(
  clean_spp == "OTCR" ~ "Snow Crab",
  clean_spp == "BTCR" ~ "Tanner Crab",
  clean_spp == "RKCR" ~ "Red King Crab",
  clean_spp == "HERR" ~ "Herring",
  TRUE                ~ "Groundfish Species"
)

# Parse selected method from params (defaults to "both" if missing)
selected_method <- if (!is.null(params$method)) trimws(tolower(params$method)) else "both"

# 2. Join estimated changes with status quo numbers using the timeline ledger tracking method
total.join.plot.annual2 <- dplyr::left_join(all_psc_status_quo_annual, total.join.plot.annual, by = c("YEAR"))
total.join.plot.annual2$est_change_method1[is.na(total.join.plot.annual2$est_change_method1)] <- 0
total.join.plot.annual2$est_change_method2[is.na(total.join.plot.annual2$est_change_method2)] <- 0

total.join.plot.annual2$net_new_PSC_m1 <- total.join.plot.annual2$sum_psc_status_quo + total.join.plot.annual2$est_change_method1
total.join.plot.annual2$net_new_PSC_m2 <- total.join.plot.annual2$sum_psc_status_quo + total.join.plot.annual2$est_change_method2
total.join.plot.annual2$LABEL          <- paste0("SPECIES")

colnames(total.join.plot.annual2) <- c("YEAR", "psc_statusquo", "psc_change_m1", "psc_change_m2", "net_new_PSC_m1", "net_new_PSC_m2", "LABEL")

# 3. STEP 1: Re-center bootstrap variance directly around the deterministic model means
total.join.plot.annual2 <- total.join.plot.annual2 %>%
  dplyr::left_join(boot_annual_ci, by = "YEAR") %>%
  dplyr::mutate(
    # Method 1 Relative Transformation
    boot_median_m1 = (lower_change_m1 + upper_change_m1) / 2,
    boot_ymin_m1   = net_new_PSC_m1 + (lower_change_m1 - boot_median_m1),
    boot_ymax_m1   = net_new_PSC_m1 + (upper_change_m1 - boot_median_m1),
    
    # Method 2 Relative Transformation
    boot_median_m2 = (lower_change_m2 + upper_change_m2) / 2,
    boot_ymin_m2   = net_new_PSC_m2 + (lower_change_m2 - boot_median_m2),
    boot_ymax_m2   = net_new_PSC_m2 + (upper_change_m2 - boot_median_m2)
  )

# 4. STEP 2: Pivot long format for clustered bar chart layout
total.join.plot.annual3 <- total.join.plot.annual2 %>% tidyr::pivot_longer(
  cols = c("psc_statusquo", "net_new_PSC_m1", "net_new_PSC_m2"),
  names_to = "psc_type",
  values_to = "psc_amount",
  values_drop_na = TRUE
)

# 5. STEP 3: Isolate upper/lower boundaries based on row categories
total.join.plot.annual3 <- total.join.plot.annual3 %>%
  dplyr::mutate(
    ymin = case_when(
      psc_type == "net_new_PSC_m1" ~ boot_ymin_m1,
      psc_type == "net_new_PSC_m2" ~ boot_ymin_m2,
      TRUE                         ~ NA_real_  # Leaves status quo clean
    ),
    ymax = case_when(
      psc_type == "net_new_PSC_m1" ~ boot_ymax_m1,
      psc_type == "net_new_PSC_m2" ~ boot_ymax_m2,
      TRUE                         ~ NA_real_  # Leaves status quo clean
    )
  )

# 6. DYNAMIC FILTERING & FACTOR MAPPING BASED ON params$method
if (selected_method == "m1") {
  total.join.plot.annual3 <- total.join.plot.annual3 %>%
    dplyr::filter(psc_type %in% c("psc_statusquo", "net_new_PSC_m1"))
  
  method_levels <- c("psc_statusquo", "net_new_PSC_m1")
  fill_values   <- c("psc_statusquo" = "#757575", "net_new_PSC_m1" = "#4daf4a")
  fill_labels   <- c("Status Quo Observed Baseline", "Method 1: Catch Volume Only (Dynamic Weekly)")

} else if (selected_method == "m2") {
  total.join.plot.annual3 <- total.join.plot.annual3 %>%
    dplyr::filter(psc_type %in% c("psc_statusquo", "net_new_PSC_m2"))
  
  method_levels <- c("psc_statusquo", "net_new_PSC_m2")
  fill_values   <- c("psc_statusquo" = "#757575", "net_new_PSC_m2" = "#e41a1c")
  fill_labels   <- c("Status Quo Observed Baseline", "Method 2: CPUE Effort Constrained (Seasonal Baseline)")

} else { # "both" or default
  method_levels <- c("psc_statusquo", "net_new_PSC_m1", "net_new_PSC_m2")
  fill_values   <- c("psc_statusquo" = "#757575", "net_new_PSC_m1" = "#4daf4a", "net_new_PSC_m2" = "#e41a1c")
  fill_labels   <- c(
    "Status Quo Observed Baseline", 
    "Method 1: Catch Volume Only (Dynamic Weekly)", 
    "Method 2: CPUE Effort Constrained (Seasonal Baseline)"
  )
}

# Enforce explicit factorization order
total.join.plot.annual3$psc_type <- factor(
  total.join.plot.annual3$psc_type,
  levels = method_levels
)

# 7. STEP 4: Render plot using dynamic matching values
ggplot(total.join.plot.annual3, aes(x = factor(YEAR), y = psc_amount, fill = psc_type)) +
  ggtitle(paste0(spp_title_fixed, ": ", title_name)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.75, alpha = 0.9) +
  geom_errorbar(
    aes(ymin = ymin, ymax = ymax, group = psc_type),
    position = position_dodge(width = 0.8),
    width = 0.25,
    color = "black",
    linewidth = 0.7,
    na.rm = TRUE
  ) +
  theme_bw() +
  labs(x = "Fishery Calendar Year", y = "Estimated Annual PSC (# Fish / Wt)") +
  scale_fill_manual(
    values = fill_values,
    labels = fill_labels,
    name   = "Closure Scenario"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    legend.position = "bottom"
  )

```

## Summary Statistics

```{r}

#| label: final-table
#| fig-width: 8
#| fig-height: 5

suppressPackageStartupMessages({
  library(dplyr)
})

# Parse selected method from params (defaults to "both" if missing)
selected_method <- if (!is.null(params$method)) trimws(tolower(params$method)) else "both"

# 1. Base table construction with all calculated metrics
final.summary.table_base <- total.join.plot.annual2 %>% 
  dplyr::select(
    Year           = YEAR, 
    PSC_Status_Quo = psc_statusquo, 
    PSC_M1_Volume  = net_new_PSC_m1, 
    Net_Change_M1  = psc_change_m1,
    PSC_M2_Effort  = net_new_PSC_m2, 
    Net_Change_M2  = psc_change_m2
  ) %>%
  dplyr::mutate(
    Pct_Change_M1 = ifelse(PSC_Status_Quo > 0, (Net_Change_M1 / PSC_Status_Quo) * 100, 0),
    Pct_Change_M2 = ifelse(PSC_Status_Quo > 0, (Net_Change_M2 / PSC_Status_Quo) * 100, 0)
  )

# 2. Dynamically select columns based on params$method
if (selected_method == "m1") {
  final.summary.table <- final.summary.table_base %>%
    dplyr::select(
      Year, 
      PSC_Status_Quo, 
      PSC_M1_Volume, 
      Net_Change_M1, 
      Pct_Change_M1
    )
} else if (selected_method == "m2") {
  final.summary.table <- final.summary.table_base %>%
    dplyr::select(
      Year, 
      PSC_Status_Quo, 
      PSC_M2_Effort, 
      Net_Change_M2, 
      Pct_Change_M2
    )
} else { # "both" or default
  final.summary.table <- final.summary.table_base %>%
    dplyr::select(
      Year, 
      PSC_Status_Quo, 
      PSC_M1_Volume, 
      Net_Change_M1, 
      Pct_Change_M1,
      PSC_M2_Effort, 
      Net_Change_M2, 
      Pct_Change_M2
    )
}

final.summary.table <- as.data.frame(final.summary.table)

# Format for clean console printing
print(dplyr::mutate_if(final.summary.table, is.numeric, round, 1))

```

## 15. Retrospective Analysis Summary

```{r}

#| label: summary-statement
#| echo: false

fmt_pct <- function(x) sprintf("%+.1f%%", x)

m1_min <- min(final.summary.table$Pct_Change_M1, na.rm = TRUE)
m1_max <- max(final.summary.table$Pct_Change_M1, na.rm = TRUE)

# Build text dynamically
if (selected_method == "m1") {
  text_out <- sprintf(
    "Under Method 1 (Catch Volume), the estimated percent change in annual PSC ranged from **%s** to **%s** relative to the Status Quo forecast.",
    fmt_pct(m1_min), fmt_pct(m1_max)
  )
} else if (selected_method == "m2") {
  m2_min <- min(final.summary.table$Pct_Change_M2, na.rm = TRUE)
  m2_max <- max(final.summary.table$Pct_Change_M2, na.rm = TRUE)
  
  text_out <- sprintf(
    "Under Method 2 (CPUE Effort), the estimated percent change in annual PSC ranged from **%s** to **%s** relative to the Status Quo forecast.",
    fmt_pct(m2_min), fmt_pct(m2_max)
  )
} else {
  m2_min <- min(final.summary.table$Pct_Change_M2, na.rm = TRUE)
  m2_max <- max(final.summary.table$Pct_Change_M2, na.rm = TRUE)
  
  text_out <- sprintf(
    "The estimated percent change in annual PSC relative to the Status Quo forecast ranged from **%s** to **%s** for Method 1 (Catch Volume) and from **%s** to **%s** for Method 2 (CPUE Effort).",
    fmt_pct(m1_min), fmt_pct(m1_max), fmt_pct(m2_min), fmt_pct(m2_max)
  )
}

# Construct Quarto Native Callout Block
callout_block <- sprintf(
  "\n\n::: {.callout-note appearance=\"simple\" icon=false}\n%s\n:::\n\n",
  text_out
)

# Force knitr to output as direct Markdown (prevents code block wrapping)
knitr::asis_output(callout_block)

```



# AR(1) PROJECTIONS
This step runs an AR(1) projection for each year using the data available to it at time, t. This produces a blind projection for each year using only data up to the previous year. In this example, the AR(1) projection of no closure (blue line) is able to stay within the ballpark of the actual observed PSC that occurred each year (black line). The AR(1) projection under a closure scenario (green line) shows only minor differences from the non-closure (blue). 

## 1. Use AR(1) to forward-project PSC catch and reductions given a closure

```{r}

# ==============================================================================
# STEP 1: DIFFERENCED ARIMA(1,1,0) WITH 5-YEAR ROLLING MEMORY & t-1 FOOTPRINT
# ==============================================================================
#| label: step1-ar1-differenced-5yr

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
})

# 1. DIFFERENCED / RANDOM WALK ARIMA(1, 1, 0) FORECASTING FUNCTION
forecast_ar1_differenced <- function(series_vec) {
  series_vec <- series_vec[!is.na(series_vec)]
  n <- length(series_vec)
  
  if (n == 0) return(c(pred = 0, se = 0))
  if (n == 1) return(c(pred = series_vec[1], se = 0))
  if (var(series_vec) == 0) return(c(pred = series_vec[n], se = 0))
  
  # If n <= 3 within the 5-year window, fallback to Last-Observation-Carried-Forward
  if (n <= 3) {
    return(c(pred = series_vec[n], se = sd(series_vec) / sqrt(pmax(1, n))))
  }
  
  fit <- tryCatch({
    suppressWarnings(
      arima(series_vec, order = c(1, 1, 0), method = "CSS-ML")
    )
  }, error = function(e) { NULL })
  
  if (!is.null(fit)) {
    p <- predict(fit, n.ahead = 1)
    return(c(pred = pmax(0, as.numeric(p$pred[1])), se = as.numeric(p$se[1])))
  } else {
    # Fallback to Last-Observation-Carried-Forward (LOCF)
    return(c(pred = series_vec[n], se = sd(series_vec, na.rm = TRUE)))
  }
}

# 2. ROLLING FORECAST LOOP WITH 5-YEAR HISTORICAL MEMORY [t-5, t-1] & t-1 FOOTPRINT
all_years  <- sort(unique(catch_data_binned$YEAR))
eval_years <- all_years[all_years >= (min(all_years) + 3)] # Evaluates 2012+

predicted_primitives_list <- list()

for (target_yr in eval_years) {
  
  prior_yr  <- target_yr - 1
  mem_start <- target_yr - 5  # STRICT 5-YEAR ROLLING MEMORY FLOOR [t-5, t-1]
  
  # A. Historical memory: strictly restricted to 5-year trailing window
  hist_memory <- catch_data_binned %>%
    dplyr::filter(
      YEAR >= mem_start & YEAR < target_yr,
      STAT_AREA %in% local_stat_areas
    ) %>%
    dplyr::mutate(
      cell_rate_mt = ifelse(GF_TOTAL_CATCH_MT >= 1.0, PSC_VALUE / GF_TOTAL_CATCH_MT, 0)
    )
  
  # B. ACTIVE SPATIAL DOMAIN: Cells active in Year t-1 ONLY
  active_cells_t1 <- catch_data_binned %>%
    dplyr::filter(YEAR == prior_yr, GF_TOTAL_CATCH_MT > 0) %>%
    dplyr::select(TIME_STEP_KEY, STAT_AREA, HARVEST_SECTOR, SPECIES_GROUP_CODE) %>%
    dplyr::distinct()
  
  # C. Fit Differenced AR(1,1,0) models & mask by t-1 footprint
  grid_forecast <- hist_memory %>%
    dplyr::group_by(TIME_STEP_KEY, STAT_AREA, HARVEST_SECTOR, SPECIES_GROUP_CODE) %>%
    dplyr::summarise(
      fc_gf   = list(forecast_ar1_differenced(GF_TOTAL_CATCH_MT)),
      fc_rate = list(forecast_ar1_differenced(cell_rate_mt)),
      fc_eff  = list(forecast_ar1_differenced(EFFORT_MINUTES)),
      .groups = "drop"
    ) %>%
    dplyr::inner_join(
      active_cells_t1,
      by = c("TIME_STEP_KEY", "STAT_AREA", "HARVEST_SECTOR", "SPECIES_GROUP_CODE")
    ) %>%
    dplyr::mutate(
      YEAR               = as.integer(target_yr),
      
      GF_TOTAL_CATCH_MT  = pmax(0, sapply(fc_gf, function(x) x["pred"])),
      se_gf_mt           = sapply(fc_gf, function(x) x["se"]),
      
      pred_rate_mt       = pmax(0, sapply(fc_rate, function(x) x["pred"])),
      
      # Derived PSC Value: GF * Rate
      PSC_VALUE          = GF_TOTAL_CATCH_MT * pred_rate_mt,
      se_psc             = sapply(fc_rate, function(x) x["se"]) * GF_TOTAL_CATCH_MT,
      
      EFFORT_MINUTES     = as.integer(round(pmax(1, sapply(fc_eff, function(x) x["pred"])))),
      se_eff_mins        = sapply(fc_eff, function(x) x["se"])
    ) %>%
    dplyr::select(
      YEAR, TIME_STEP_KEY, STAT_AREA, HARVEST_SECTOR, SPECIES_GROUP_CODE,
      GF_TOTAL_CATCH_MT, PSC_VALUE, EFFORT_MINUTES,
      se_gf_mt, se_psc, se_eff_mins
    )
  
  predicted_primitives_list[[as.character(target_yr)]] <- grid_forecast
}

predicted_primitives_df <- do.call(rbind, predicted_primitives_list)

```


## 2. Run each estimation method

```{r}
# ==============================================================================
# STEP 2: RUN METHOD 1 (PSC/GF RATES) & METHOD 2 (CPUE EFFORT)
# ==============================================================================
#| label: step2-pipeline-both-scenarios

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
})

# 1. CALCULATE EMPIRICAL RATE CV DIRECTLY FROM OBSERVED DATASET
empirical_rate_cv <- catch_data_binned %>%
  dplyr::filter(!(STAT_AREA %in% closure.stat.areas)) %>%
  dplyr::group_by(YEAR, HARVEST_SECTOR) %>%
  dplyr::summarise(
    annual_gf_mt  = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE),
    annual_psc    = sum(PSC_VALUE, na.rm = TRUE),
    annual_rate   = ifelse(annual_gf_mt > 0, annual_psc / annual_gf_mt, 0),
    .groups = "drop"
  ) %>%
  dplyr::group_by(HARVEST_SECTOR) %>%
  dplyr::summarise(
    cv_data = ifelse(mean(annual_rate, na.rm = TRUE) > 0, 
                     sd(annual_rate, na.rm = TRUE) / mean(annual_rate, na.rm = TRUE), 
                     0.05),
    .groups = "drop"
  )

cv_data_factor <- mean(empirical_rate_cv$cv_data, na.rm = TRUE)

# 2. CORE REDISTRIBUTION PIPELINE
run_redistribution_pipeline <- function(pred_df, is_closure_active) {
  stat_areas_closed <- as.character(closure.stat.areas)
  
  pred_impacted_cluster <- pred_df %>%
    dplyr::mutate(
      STAT_AREA = stringr::str_pad(trimws(as.character(STAT_AREA)), width = 6, side = "left", pad = "0"),
      inside    = ifelse(STAT_AREA %in% stat_areas_closed, "yes", "no"),
      impacted  = ifelse(is_closure_active, "yes", "no")
    )
  
  # --- STEP A: INSIDE CATCH SUMMARY ---
  gf_inside_cluster_pred <- pred_impacted_cluster %>%
    dplyr::filter(inside == "yes") %>%
    dplyr::group_by(YEAR, TIME_STEP_KEY, HARVEST_SECTOR) %>%
    dplyr::summarise(
      tons_gf_inside_wk     = sum(GF_TOTAL_CATCH_MT * (impacted == "yes"), na.rm = TRUE),
      psc_measure_inside_wk = sum(PSC_VALUE * (impacted == "no"), na.rm = TRUE),
      # Safely handle NA effort for unobserved landings
      effort_inside_wk      = ifelse(all(is.na(EFFORT_MINUTES)), NA_real_, sum(EFFORT_MINUTES, na.rm = TRUE)),
      .groups = "drop"
    )
  
  # --- STEP B: OUTSIDE PERFORMANCE MATRIX ---
  gf_outside_cluster_pred <- pred_impacted_cluster %>%
    dplyr::filter(inside == "no") %>%
    dplyr::mutate(DYNAMIC_SEASONAL_KEY = TIME_STEP_KEY)
  
  # 3-Year Rolling Historical Baseline Profile (t-3 to t-1 relative to target_yr)
  rolling_cpue_list_pred <- list()
  all_pred_years <- sort(unique(gf_outside_cluster_pred$YEAR))
  min_pred_yr    <- min(all_pred_years)
  
  for (target_yr in all_pred_years) {
    w_start <- max(min_pred_yr, target_yr - 3)
    w_end   <- max(min_pred_yr, target_yr - 1)
    
    if(target_yr == min_pred_yr) { w_start <- target_yr; w_end <- target_yr }
    
    hist_window <- catch_data_binned %>%
      dplyr::filter(
        !(STAT_AREA %in% stat_areas_closed),
        YEAR >= w_start & YEAR <= w_end
      ) %>%
      dplyr::group_by(TIME_STEP_KEY, STAT_AREA, HARVEST_SECTOR) %>%
      dplyr::summarise(
        sum_tons_3yr   = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE),
        sum_effort_3yr = ifelse(all(is.na(EFFORT_MINUTES)), NA_real_, sum(EFFORT_MINUTES, na.rm = TRUE)),
        sum_psc_3yr    = sum(PSC_VALUE, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      dplyr::mutate(
        hist_gf_cpue_seasonal  = ifelse(target_yr > min_pred_yr & !is.na(sum_effort_3yr) & sum_effort_3yr > 0, 
                                        sum_tons_3yr / sum_effort_3yr, NA_real_),
        hist_psc_rate_mins_3yr = ifelse(target_yr > min_pred_yr & !is.na(sum_effort_3yr) & sum_effort_3yr > 0, 
                                        sum_psc_3yr / sum_effort_3yr, NA_real_),
        YEAR = target_yr,
        DYNAMIC_SEASONAL_KEY = TIME_STEP_KEY
      ) %>%
      dplyr::select(YEAR, DYNAMIC_SEASONAL_KEY, STAT_AREA, HARVEST_SECTOR, hist_gf_cpue_seasonal, hist_psc_rate_mins_3yr)
    
    rolling_cpue_list_pred[[as.character(target_yr)]] <- hist_window
  }
  
  historical_cpue_profile_pred <- do.call(rbind, rolling_cpue_list_pred)
  
  # Regional Sector Fallback Grid (Separated cleanly by HARVEST_SECTOR)
  reg_fallbacks_pred <- catch_data_binned %>%
    dplyr::filter(!(STAT_AREA %in% closure.stat.areas)) %>%
    dplyr::group_by(TIME_STEP_KEY, HARVEST_SECTOR) %>%
    dplyr::summarise(
      reg_cpue_fallback     = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE) / sum(EFFORT_MINUTES, na.rm = TRUE),
      reg_psc_mins_fallback = sum(PSC_VALUE, na.rm = TRUE) / sum(EFFORT_MINUTES, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::rename(DYNAMIC_SEASONAL_KEY = TIME_STEP_KEY)
  
  weekly_base_metrics_pred <- gf_outside_cluster_pred %>%
    dplyr::group_by(YEAR, TIME_STEP_KEY, DYNAMIC_SEASONAL_KEY, STAT_AREA, HARVEST_SECTOR) %>%
    dplyr::summarise(
      tons_gf_wk           = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE),
      total_effort_minutes = ifelse(all(is.na(EFFORT_MINUTES)), NA_real_, sum(EFFORT_MINUTES, na.rm = TRUE)),
      total_psc_count      = sum(PSC_VALUE, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      mean_psc_rate_wk = ifelse(tons_gf_wk > 0, total_psc_count / tons_gf_wk, 0)
    ) %>%
    dplyr::left_join(
      gf_outside_cluster_pred %>%
        dplyr::group_by(YEAR, TIME_STEP_KEY, HARVEST_SECTOR) %>%
        dplyr::summarise(tons_gf_annual = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE), .groups = "drop"),
      by = c("YEAR", "TIME_STEP_KEY", "HARVEST_SECTOR")
    ) %>%
    dplyr::left_join(
      gf_outside_cluster_pred %>%
        dplyr::group_by(YEAR, TIME_STEP_KEY, STAT_AREA, HARVEST_SECTOR) %>%
        dplyr::summarise(tons_gf_annual_stat = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE), .groups = "drop"),
      by = c("YEAR", "TIME_STEP_KEY", "STAT_AREA", "HARVEST_SECTOR")
    ) %>%
    dplyr::mutate(
      proportion_outside_wk = ifelse(tons_gf_annual > 0, tons_gf_annual_stat / tons_gf_annual, 0)
    )
  
  sum.stat.timeseries.gf_pred <- weekly_base_metrics_pred %>%
    dplyr::left_join(
      historical_cpue_profile_pred,
      by = c("YEAR", "DYNAMIC_SEASONAL_KEY", "STAT_AREA", "HARVEST_SECTOR")
    ) %>%
    dplyr::left_join(
      reg_fallbacks_pred,
      by = c("DYNAMIC_SEASONAL_KEY", "HARVEST_SECTOR")
    ) %>%
    dplyr::mutate(
      # Coalesce to regional fallback whenever local stat cell CPUE is NA
      hist_gf_cpue_seasonal  = dplyr::coalesce(hist_gf_cpue_seasonal, reg_cpue_fallback),
      hist_psc_rate_mins_3yr = dplyr::coalesce(hist_psc_rate_mins_3yr, reg_psc_mins_fallback, 0)
    ) %>%
    as.data.frame()
  
  # --- STEP C: EXECUTE DISPLACEMENT MATRIX ---
  total.join.master.pred <- sum.stat.timeseries.gf_pred %>%
    dplyr::left_join(
      gf_inside_cluster_pred %>% dplyr::select(YEAR, TIME_STEP_KEY, HARVEST_SECTOR, tons_gf_inside_wk),
      by = c("YEAR", "TIME_STEP_KEY", "HARVEST_SECTOR")
    ) %>%
    dplyr::mutate(
      tons_gf_inside_wk = tidyr::replace_na(tons_gf_inside_wk, 0),
      gf_catch_new      = tons_gf_inside_wk * proportion_outside_wk,
      
      # Method 1
      new_PSC_method1   = gf_catch_new * mean_psc_rate_wk,
      
      # Method 2
      effort_minutes_needed = ifelse(!is.na(hist_gf_cpue_seasonal) & hist_gf_cpue_seasonal > 0, 
                                     gf_catch_new / hist_gf_cpue_seasonal, 0),
      new_PSC_method2       = effort_minutes_needed * tidyr::replace_na(hist_psc_rate_mins_3yr, 0)
    )
  
  ledger_weekly_pred <- total.join.master.pred %>%
    dplyr::group_by(YEAR, TIME_STEP_KEY, HARVEST_SECTOR) %>%
    dplyr::summarise(
      wk_out_base_psc = sum(total_psc_count, na.rm = TRUE),
      new_PSC_m1_wk   = sum(new_PSC_method1, na.rm = TRUE),
      new_PSC_m2_wk   = sum(new_PSC_method2, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::left_join(
      gf_inside_cluster_pred %>% dplyr::select(YEAR, TIME_STEP_KEY, HARVEST_SECTOR, psc_measure_inside_wk),
      by = c("YEAR", "TIME_STEP_KEY", "HARVEST_SECTOR")
    ) %>%
    dplyr::mutate(
      psc_measure_inside_wk = tidyr::replace_na(psc_measure_inside_wk, 0),
      net_total_m1_wk       = wk_out_base_psc + psc_measure_inside_wk + new_PSC_m1_wk,
      net_total_m2_wk       = wk_out_base_psc + psc_measure_inside_wk + new_PSC_m2_wk
    )
  
  total.join.plot.annual.pred <- ledger_weekly_pred %>%
    dplyr::group_by(YEAR) %>%
    dplyr::summarise(
      proj_m1_mean = sum(net_total_m1_wk, na.rm = TRUE),
      proj_m2_mean = sum(net_total_m2_wk, na.rm = TRUE),
      
      proj_m1_se   = sqrt(sum((net_total_m1_wk * cv_data_factor)^2, na.rm = TRUE)),
      proj_m2_se   = sqrt(sum((net_total_m2_wk * cv_data_factor)^2, na.rm = TRUE)),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      m1_lower = pmax(0, proj_m1_mean - 1.96 * proj_m1_se),
      m1_upper = proj_m1_mean + 1.96 * proj_m1_se,
      m2_lower = pmax(0, proj_m2_mean - 1.96 * proj_m2_se),
      m2_upper = proj_m2_mean + 1.96 * proj_m2_se
    )
  
  return(total.join.plot.annual.pred)
}

# Run both scenario passes
pred_sq_df      <- run_redistribution_pipeline(predicted_primitives_df, is_closure_active = FALSE)
pred_closure_df <- run_redistribution_pipeline(predicted_primitives_df, is_closure_active = TRUE)

```



## 3. Plot against observations

```{r}

#| label: step3-validation-plot-all5
#| fig-width: 11
#| fig-height: 7

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(scales)
})

# Parse selected method from params (defaults to "both" if missing)
selected_method <- if (!is.null(params$method)) trimws(tolower(params$method)) else "both"

# ==============================================================================
# 1. JOIN OBSERVED REALITY + STATUS QUO FORECAST + CLOSURE FORECASTS
# ==============================================================================
val_all5_df <- total.join.plot.annual2 %>%
  dplyr::select(YEAR, psc_statusquo) %>%
  dplyr::inner_join(
    pred_sq_df %>% dplyr::select(YEAR, sq_ar1_mean = proj_m1_mean, sq_ar1_lower = m1_lower, sq_ar1_upper = m1_upper),
    by = "YEAR"
  ) %>%
  dplyr::inner_join(
    pred_closure_df %>% dplyr::select(YEAR, proj_m1_mean, m1_lower, m1_upper, proj_m2_mean, m2_lower, m2_upper),
    by = "YEAR"
  )

# Base Line Series
plot_lines_base <- val_all5_df %>%
  dplyr::select(YEAR, psc_statusquo, sq_ar1_mean, proj_m1_mean, proj_m2_mean)

# Base Ribbon Series
plot_ribbons_sq <- val_all5_df %>%
  dplyr::select(YEAR, sq_ar1_lower, sq_ar1_upper) %>%
  dplyr::mutate(Metric = "sq_ar1_mean", ymin = sq_ar1_lower, ymax = sq_ar1_upper) %>%
  dplyr::select(YEAR, Metric, ymin, ymax)

plot_ribbons_m1 <- val_all5_df %>%
  dplyr::select(YEAR, m1_lower, m1_upper) %>%
  dplyr::mutate(Metric = "proj_m1_mean", ymin = m1_lower, ymax = m1_upper) %>%
  dplyr::select(YEAR, Metric, ymin, ymax)

plot_ribbons_m2 <- val_all5_df %>%
  dplyr::select(YEAR, m2_lower, m2_upper) %>%
  dplyr::mutate(Metric = "proj_m2_mean", ymin = m2_lower, ymax = m2_upper) %>%
  dplyr::select(YEAR, Metric, ymin, ymax)

# Filter columns, ribbons, and scales based on selected_method
if (selected_method == "m1") {
  plot_lines <- plot_lines_base %>% dplyr::select(-proj_m2_mean)
  plot_ribbons <- dplyr::bind_rows(plot_ribbons_sq, plot_ribbons_m1)
  
  metric_levels <- c("psc_statusquo", "sq_ar1_mean", "proj_m1_mean")
  metric_labels <- c(
    "Observed Status Quo (Baseline Reality)",
    "AR(1) Unconditional Forecast (Status Quo Validation)",
    "Method 1: Projected Catch Volume Model (Closure Active)"
  )
  
  color_vals    <- c("#111111", "#377eb8", "#4daf4a")
  fill_vals     <- c("transparent", "#377eb8", "#4daf4a")
  linetype_vals <- c("solid", "dashed", "dashed")
  shape_vals    <- c(16, 17, 15)
  subtitle_text <- "AR(1) Status Quo Validation and Catch Volume Model (Method 1) vs Observed Reality"

} else if (selected_method == "m2") {
  plot_lines <- plot_lines_base %>% dplyr::select(-proj_m1_mean)
  plot_ribbons <- dplyr::bind_rows(plot_ribbons_sq, plot_ribbons_m2)
  
  metric_levels <- c("psc_statusquo", "sq_ar1_mean", "proj_m2_mean")
  metric_labels <- c(
    "Observed Status Quo (Baseline Reality)",
    "AR(1) Unconditional Forecast (Status Quo Validation)",
    "Method 2: Projected CPUE Effort Model (Closure Active)"
  )
  
  color_vals    <- c("#111111", "#377eb8", "#e41a1c")
  fill_vals     <- c("transparent", "#377eb8", "#e41a1c")
  linetype_vals <- c("solid", "dashed", "dotdash")
  shape_vals    <- c(16, 17, 18)
  subtitle_text <- "AR(1) Status Quo Validation and CPUE Effort Model (Method 2) vs Observed Reality"

} else { # "both" or default fallback
  plot_lines <- plot_lines_base
  plot_ribbons <- dplyr::bind_rows(plot_ribbons_sq, plot_ribbons_m1, plot_ribbons_m2)
  
  metric_levels <- c("psc_statusquo", "sq_ar1_mean", "proj_m1_mean", "proj_m2_mean")
  metric_labels <- c(
    "Observed Status Quo (Baseline Reality)",
    "AR(1) Unconditional Forecast (Status Quo Validation)",
    "Method 1: Projected Catch Volume Model (Closure Active)",
    "Method 2: Projected CPUE Effort Model (Closure Active)"
  )
  
  color_vals    <- c("#111111", "#377eb8", "#4daf4a", "#e41a1c")
  fill_vals     <- c("transparent", "#377eb8", "#4daf4a", "#e41a1c")
  linetype_vals <- c("solid", "dashed", "dashed", "dotdash")
  shape_vals    <- c(16, 17, 15, 18)
  subtitle_text <- "AR(1) Status Quo Validation and Closure Redistribution Models (Methods 1 & 2) vs Observed Reality"
}

# Pivot active lines to long format
plot_lines_long <- plot_lines %>%
  tidyr::pivot_longer(-YEAR, names_to = "Metric", values_to = "Mean_Value")

# Assign names to scale vectors for exact manual legend matching
names(color_vals)    <- metric_labels
names(fill_vals)     <- metric_labels
names(linetype_vals) <- metric_labels
names(shape_vals)    <- metric_labels

# Combine lines and ribbons with strict factor ordering
plot_unified_data <- dplyr::left_join(plot_lines_long, plot_ribbons, by = c("YEAR", "Metric")) %>%
  dplyr::mutate(
    Metric = factor(Metric, levels = metric_levels, labels = metric_labels)
  )

# ==============================================================================
# 2. RENDER THE COMPLETE MASTER CHART
# ==============================================================================
ggplot(plot_unified_data, aes(x = YEAR, y = Mean_Value, color = Metric, fill = Metric, linetype = Metric, shape = Metric)) +
  geom_ribbon(aes(ymin = ymin, ymax = ymax), alpha = 0.12, color = NA) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = sort(unique(plot_unified_data$YEAR))) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_manual(values = color_vals, name = "Scenario Model") +
  scale_fill_manual(values = fill_vals, name = "Scenario Model") +
  scale_linetype_manual(values = linetype_vals, name = "Scenario Model") +
  scale_shape_manual(values = shape_vals, name = "Scenario Model") +
  theme_bw() +
  labs(
    title = paste0("Unified AR(1) Forward Projection & Scenario Validation: ", title_name),
    subtitle = subtitle_text,
    x = "Fishery Calendar Year",
    y = "Total Regional PSC (# / MT)"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    plot.subtitle = element_text(face = "italic", size = 10, hjust = 0.5),
    legend.position = "bottom",
    legend.direction = "vertical",
    axis.text.x = element_text(face = "bold", size = 10),
    panel.grid.minor = element_blank()
  )
```

## 4. Summarize PSC Change

```{r}

#| label: final-table2

suppressPackageStartupMessages({
  library(dplyr)
  library(knitr)
})

# Parse selected method from params (defaults to "both" if missing)
selected_method <- if (!is.null(params$method)) trimws(tolower(params$method)) else "both"

# ==============================================================================
# SUMMARY TABLE: PSC PROJECTED TOTALS & PERCENT CHANGES
# ==============================================================================

# 1. Compute full metrics table
final.summary.table_base <- val_all5_df %>% 
  dplyr::mutate(
    Net_Change_M1 = proj_m1_mean - sq_ar1_mean,
    Net_Change_M2 = proj_m2_mean - sq_ar1_mean,
    Pct_Change_M1 = ifelse(sq_ar1_mean > 0, (Net_Change_M1 / sq_ar1_mean) * 100, 0),
    Pct_Change_M2 = ifelse(sq_ar1_mean > 0, (Net_Change_M2 / sq_ar1_mean) * 100, 0)
  )

# 2. Dynamically filter columns based on params$method
if (selected_method == "m1") {
  final.summary.table2 <- final.summary.table_base %>%
    dplyr::select(
      Year           = YEAR, 
      PSC_Status_Quo = sq_ar1_mean, 
      PSC_M1_Volume  = proj_m1_mean, 
      Net_Change_M1, 
      Pct_Change_M1
    )
} else if (selected_method == "m2") {
  final.summary.table2 <- final.summary.table_base %>%
    dplyr::select(
      Year           = YEAR, 
      PSC_Status_Quo = sq_ar1_mean, 
      PSC_M2_Effort  = proj_m2_mean, 
      Net_Change_M2, 
      Pct_Change_M2
    )
} else { # "both" or default
  final.summary.table2 <- final.summary.table_base %>%
    dplyr::select(
      Year           = YEAR, 
      PSC_Status_Quo = sq_ar1_mean, 
      PSC_M1_Volume  = proj_m1_mean, 
      Net_Change_M1, 
      Pct_Change_M1,
      PSC_M2_Effort  = proj_m2_mean, 
      Net_Change_M2, 
      Pct_Change_M2
    )
}

final.summary.table2 <- as.data.frame(final.summary.table2)

# Console Print (Clean rounded numbers)
print(dplyr::mutate_if(final.summary.table2, is.numeric, round, 1))

```

## 5. AR(1) Summary

```{r}

#| label: summary-statement
#| echo: false

fmt_pct <- function(x) sprintf("%+.1f%%", x)

m1_min <- min(final.summary.table2$Pct_Change_M1, na.rm = TRUE)
m1_max <- max(final.summary.table2$Pct_Change_M1, na.rm = TRUE)

# Build text dynamically
if (selected_method == "m1") {
  text_out <- sprintf(
    "Under Method 1 (Catch Volume), the estimated percent change in annual PSC ranged from **%s** to **%s** relative to the Status Quo forecast.",
    fmt_pct(m1_min), fmt_pct(m1_max)
  )
} else if (selected_method == "m2") {
  m2_min <- min(final.summary.table2$Pct_Change_M2, na.rm = TRUE)
  m2_max <- max(final.summary.table2$Pct_Change_M2, na.rm = TRUE)
  
  text_out <- sprintf(
    "Under Method 2 (CPUE Effort), the estimated percent change in annual PSC ranged from **%s** to **%s** relative to the Status Quo forecast.",
    fmt_pct(m2_min), fmt_pct(m2_max)
  )
} else {
  m2_min <- min(final.summary.table2$Pct_Change_M2, na.rm = TRUE)
  m2_max <- max(final.summary.table2$Pct_Change_M2, na.rm = TRUE)
  
  text_out <- sprintf(
    "The estimated percent change in annual PSC relative to the Status Quo forecast ranged from **%s** to **%s** for Method 1 (Catch Volume) and from **%s** to **%s** for Method 2 (CPUE Effort).",
    fmt_pct(m1_min), fmt_pct(m1_max), fmt_pct(m2_min), fmt_pct(m2_max)
  )
}

# Construct Quarto Native Callout Block
callout_block <- sprintf(
  "\n\n::: {.callout-note appearance=\"simple\" icon=false}\n%s\n:::\n\n",
  text_out
)

# Force knitr to output as direct Markdown (prevents code block wrapping)
knitr::asis_output(callout_block)

```

