---
title: "Proportional Groundfish Redistribution"
author: "Mason Smith (mason.smith@noaa.gov)"
date: today
format: 
  html:
    toc: true
    code-fold: true
    self-contained: true
params:
  time_step: "weekly" # Options: "weekly", "monthly", "annual"
  timeframe: [2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024, 2025]
  psc_spp: RKCR
  closure: rkcssa
  buffer: 100000
execute:
  warning: false
  message: false
---

## 1. Setup

This analysis predicts the spatial movement and net PSC change for displaced vessels under hypothetical closures of the red king crab sub savings area (RKCSSA) and validates predictions against closures in 2022 and 2023. 

```{r}
#| label: setup

#| label: setup
#| include: false

# Load packages
library(dplyr)
library(terra)
library(ggplot2)
library(lubridate)
library(tidyr)
library(units)
library(gridExtra)
library(tidyterra)
library(ggpubr)
library(patchwork)
library(ggpattern)
library(data.table)
library(sf)

options(scipen=999)

```


## 2. Set Titles

Assign titles, timeframe, and closures

```{r}

spp_title = case_when(
      params$psc_spp == "OTCR"  ~ "Snow Crab",
      params$psc_spp == "BTCR"  ~ "Tanner Crab",
      params$psc_spp == "RKCR"  ~ "Red King Crab",
      params$psc_spp == "HERR"  ~ "Herring",
      TRUE                          ~ NA # Default: NA string
    )


title_name = case_when(
      params$closure == "smihca"  ~ "Saint Matthews HCA",
      params$closure == "nunivak"  ~ "Nunivak HCA",
      params$closure == "rkcssa"  ~ "Red King Crab Sub Savings Area",
      TRUE                          ~ NA # Default: NA string
    )

```



## 3. Spatial Data and Grid Setup

Loading the Alaska background and statistical area shapefiles to create the time-series grid.

```{r}

#| label: load-spatial
stat_areas <- terra::vect("shapefiles/State_Stat_2004.shp")

# Ensure it stays a character string and strip any accidental whitespace
stat_areas$STAT_AREA <- trimws(as.character(stat_areas$STAT_AREA))

# Create timeseries list
stat.list <- list()
for (i in params$timeframe){
  stat_shape <- stat_areas
  stat_shape$YEAR <- i
  stat.list[[as.character(i)]] <- stat_shape
}
stat.areas.timeseries <- terra::vect(stat.list)


```

## 4. Closure Assignment

Define the closure area based on a polygon. This step uses an overlay threshold of 10% to determine if a closure includes a stat_area, rather than just touching it. This step also establishes the surrounding stat_areas to use for the analysis, based on a buffer parameter chosen (here, 100 km).

```{r}

#| label: closure-setup

# 1. Load the polygons using terra
closure_polygon       <- terra::vect(paste0("shapefiles/", params$closure, ".shp"))
stat_areas_with_area  <- terra::vect("shapefiles/State_Stat_2004.shp") 

# Ensure character type right away and calculate individual cell base areas
stat_areas_with_area$STAT_AREA <- trimws(as.character(stat_areas_with_area$STAT_AREA))
stat_areas_with_area$cell_base_area <- terra::expanse(stat_areas_with_area)

# 2. Project stat areas if they don't match the closure polygon CRS
if (crs(stat_areas_with_area) != crs(closure_polygon)) {
  stat_areas_with_area <- terra::project(stat_areas_with_area, crs(closure_polygon))
}

# 3. Build a geometric area overlap intersection
intersected_geometry <- terra::intersect(stat_areas_with_area, closure_polygon)

# 4. FIXED: Calculate what % of the LOCAL STAT AREA CELL is covered by the closure
intersected_geometry$overlap_area <- terra::expanse(intersected_geometry)
intersected_geometry$pct_of_cell_covered <- (intersected_geometry$overlap_area / intersected_geometry$cell_base_area) * 100

# 5. Extract valid stat areas based on local coverage 
# (Using 1.0% here catches any cell where the closure cuts through a meaningful chunk)
threshold_pct       <- 10.0
valid_intersections <- as.data.frame(intersected_geometry) %>% 
  filter(pct_of_cell_covered > threshold_pct)

valid_stat_areas   <- unique(as.character(valid_intersections$STAT_AREA))
closure.stat.areas <- trimws(valid_stat_areas)

# 6. Separate the dataset into closure and non-closure polygons
is_overlapping <- stat_areas_with_area$STAT_AREA %in% valid_stat_areas

# Isolate statistical areas meeting the threshold requirements
closure        <- stat_areas_with_area[is_overlapping, ]
closure$closed <- "yes"
closure2       <- terra::aggregate(closure)

# Isolate statistical areas that fell below the threshold requirements
non_closure        <- stat_areas_with_area[!is_overlapping, ]
non_closure$closed <- "no"

# ==============================================================================
# Generate the 100 km spatial buffer around the true aggregated closure
# ==============================================================================
closure_buffer <- terra::buffer(closure2, width = params$buffer) 

# Identify which stat areas fall within this 100 km window
local_stat_matches <- terra::relate(stat_areas_with_area, closure_buffer, "intersects")
local_stat_areas   <- stat_areas_with_area$STAT_AREA[as.vector(local_stat_matches)]

# 7. Quick visual validation check
plot(stat_areas_with_area[stat_areas_with_area$STAT_AREA %in% local_stat_areas, ], border="gray80")
plot(closure, border = "blue", lwd = 2, add = TRUE)
plot(closure_polygon, border = "red", lwd = 1.5, add = TRUE)
plot(closure_buffer, border = "purple", lty = 2, lwd = 1.5, add = TRUE)

```

## 5. Catch Data

Load catch data from the catch accounting system (CAS) and assign to stat_areas.

```{r}

#| label: catch-data

# 1. Read your pre-packaged standalone species file 
catch_data_raw <- read.csv(paste0("data/catch_data_", params$psc_spp, ".csv"), header = TRUE)
colnames(catch_data_raw) <- toupper(colnames(catch_data_raw))

# 2. Slice out data matching parameters and dynamically bin dates based on YAML parameter
catch_data <- catch_data_raw %>%
  dplyr::filter(
    YEAR %in% params$timeframe,
    SPECIES_GROUP_CODE == params$psc_spp,
    PROCESSING_SECTOR == "CP"
  ) %>%
  mutate(
    STAT_AREA = stringr::str_pad(trimws(as.character(STAT_AREA)), width = 6, side = "left", pad = "0"),
    
    # Force clean date classes
    CATCH_ACTIVITY_DATE = as.Date(CATCH_ACTIVITY_DATE),
    WEEK_END_DATE       = as.Date(WEEK_END_DATE),
    
    # === THE DYNAMIC TIME-STEP SWITCH ===
    # This assigns a universal grouping key based on what you chose at the top of the file
    TIME_STEP_KEY = case_when(
      params$time_step == "annual"  ~ as.character(YEAR),
      params$time_step == "monthly" ~ format(CATCH_ACTIVITY_DATE, "%Y-%m"),
      TRUE                          ~ as.character(WEEK_END_DATE) # Default: weekly string
    )
  )

# 3. Collapse multiple daily entries into the chosen time step *before* doing spatial math
# This prevents inflating weights when changing time horizons
catch_data_binned <- catch_data %>%
  group_by(YEAR, TIME_STEP_KEY, STAT_AREA, PROCESSING_SECTOR, SPECIES_GROUP_CODE) %>%
  summarise(
    # Aggregate to prevent duplicating values inside identical spatial blocks
    GF_TOTAL_CATCH_MT = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE),
    PSC_VALUE         = sum(PSC_VALUE, na.rm = TRUE),
    
    # Pass a single baseline date along for Section 7's closure timeline comparison
    CATCH_ACTIVITY_DATE = min(CATCH_ACTIVITY_DATE),
    .groups = "drop"
  )

# 4. Align the master spatial object and merge
stat_areas_with_area$STAT_AREA <- stringr::str_pad(trimws(as.character(stat_areas_with_area$STAT_AREA)), width = 6, side = "left", pad = "0")

# Merge catch data with spatial dataframe
catch_data_full <- merge(stat_areas_with_area, catch_data_binned, by = "STAT_AREA", all.x = FALSE)

# Subset master catch data strictly to your 100km buffer zone
catch_data2 <- subset(catch_data_full, catch_data_full$STAT_AREA %in% local_stat_areas)

```

## 6. Visualizations

Example of weekly groundfish catch.

```{r}

#| label: fig-gf-catch
#| fig-cap: "Simulated weekly groundfish catch distributions."
#| fig-width: 12
#| fig-height: 8
ggplot() +
  tidyterra::geom_spatvector(data = subset(catch_data2, catch_data2$YEAR==2018), aes(fill = GF_TOTAL_CATCH_MT), 
                             alpha = 0.8, col = NA, na.rm = TRUE) +
  scale_fill_distiller(palette = "GnBu", name = "Catch (MT)") +
  tidyterra::geom_spatvector(data = closure2, aes(color = "Closure"), 
                             fill = NA, alpha = 0.5, lwd = 0.5) +
  scale_color_manual(values = c("Closure" = "red"), name = NULL) +
  facet_wrap(~TIME_STEP_KEY, ncol=10) +
  theme_bw() + theme(plot.title = element_text(hjust = 0.5), axis.title = element_blank())

```

Example of weekly PSC catch.

```{r}

#| label: fig-psc-catch
#| fig-cap: "Simulated weekly PSC distributions."
#| fig-width: 12
#| fig-height: 8

ggplot() +
  tidyterra::geom_spatvector(data = subset(catch_data2, catch_data2$YEAR==2018), aes(fill = PSC_VALUE), 
                             alpha = 0.8, col = NA, na.rm = TRUE) +
  scale_fill_distiller(palette = "GnBu", name = "PSC Catch (Number)") +
  tidyterra::geom_spatvector(data = closure2, aes(color = "Closure"), 
                             fill = NA, alpha = 0.5, lwd = 0.5) +
  scale_color_manual(values = c("Closure" = "red"), name = NULL) +
  facet_wrap(~TIME_STEP_KEY, ncol=10) +
  theme_bw() + theme(plot.title = element_text(hjust = 0.5), axis.title = element_blank())

```



## 7. Assign Closure Dates & Isolate Impacted Hauls

For full-year closures, January 1 was chosen.

```{r}

#| label: impacted-hauls
#| echo: false

#sum status quo PSC by year for later joins
all_psc_status_quo_annual <- as.data.frame(catch_data2) %>%
  group_by(YEAR) %>%
  reframe(
    sum_psc_status_quo = sum(PSC_VALUE)
  )

#assign dates to data ---
# Automatically generate January 1st for every year in your global parameter timeline
closure_dates <- data.frame(
  YEAR = params$timeframe,
  closure_date = paste0(params$timeframe, "-01-01")
)

# Ensure the new column is explicitly treated as a clean Date class
closure_dates$closure_date <- as.Date(closure_dates$closure_date)

#assign dates (Joined strictly by YEAR so it applies to all processing sectors)
all_psc_closures_dates <- as.data.frame(dplyr::left_join(catch_data2, closure_dates, by=c("YEAR")))


#assign groundfish and psc based on date ---
all_psc_closures_dates$closure <- ifelse(!is.na(all_psc_closures_dates$closure_date), "yes", "no")
all_psc_closures_dates$impacted <- ifelse(all_psc_closures_dates$CATCH_ACTIVITY_DATE >= all_psc_closures_dates$closure_date, "yes", "no")
all_psc_closures_dates$tons_gf_after_closure <- ifelse(all_psc_closures_dates$impacted=="yes", all_psc_closures_dates$GF_TOTAL_CATCH_MT, 0)
all_psc_closures_dates$tons_gf_no_closure <- ifelse(all_psc_closures_dates$impacted=="no" | is.na(all_psc_closures_dates$impacted), all_psc_closures_dates$GF_TOTAL_CATCH_MT, 0)


#isolate impacted hauls
all_psc_impacted <- subset(all_psc_closures_dates, impacted=="yes")

# ==============================================================================
# Final Ledger Prep: Create ledger_df_clean for validation blocks
# ==============================================================================

ledger_df_clean <- as.data.frame(all_psc_closures_dates) %>%
  dplyr::mutate(
    # Ensure spatial codes are clean 6-character strings
    STAT_AREA = stringr::str_pad(trimws(as.character(STAT_AREA)), width = 6, side = "left", pad = "0"),
    # Identify if a row belongs inside the active regulatory box
    inside    = ifelse(STAT_AREA %in% closure.stat.areas, "yes", "no")
  ) %>%
  # Restrict evaluation strictly to your 100km buffer footprint outside the closure
  dplyr::filter(STAT_AREA %in% local_stat_areas, inside == "no")

```

## 8. Calculate Catch and Proportions Inside/Outside Closure

This step summarizes the data used in the main calculations - the proportions of outside groundish catch where displaced catch is applied to, under the assumption that the fishery will meet the same overall total catch, the total catch inside the closure to be applied to the proportions, the new PSC as a result of the new GF catch applied to local PSC rates, and the PSC caught inside the closure that will be subtracted from the net total.

```{r}

#| label: rates-props
#| echo: false

# Identify Inside/Outside after a closure would have occurred  ---

# Ensure your closed area matching vector is treated as clean characters
stat_areas_closed <- as.character(closure.stat.areas) 

# Identify which hauls occurred inside/outside the closed area
all_psc_impacted_cluster <- all_psc_impacted %>%
  # Safe character-to-character array matching
  mutate(inside = case_when(trimws(as.character(STAT_AREA)) %in% stat_areas_closed ~ "yes", TRUE ~ NA_character_))

all_psc_impacted_cluster$inside[is.na(all_psc_impacted_cluster$inside)] <- "no" 


# -- inside catch -- #
gf_inside_cluster <- subset(all_psc_impacted_cluster, inside == "yes")

gf_inside_cluster2 <- gf_inside_cluster %>%
  group_by(YEAR, TIME_STEP_KEY, PROCESSING_SECTOR) %>%
  summarise(
    tons_gf_inside_wk = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE),
    psc_measure_inside_wk = sum(PSC_VALUE, na.rm = TRUE),
    .groups = "drop"
  )

# -- outside catch -- #
gf_outside_cluster <- subset(all_psc_impacted_cluster, inside == "no")

gf_outside_cluster2 <- gf_outside_cluster %>%
  group_by(YEAR, TIME_STEP_KEY, PROCESSING_SECTOR) %>%
  summarise(
    tons_gf_outside_wk = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE),
    psc_measure_outside_wk = sum(PSC_VALUE, na.rm = TRUE),
    .groups = "drop"
  )


# summarize mean psc rates and calculate ratio of means  ---
sum.stat.timeseries.gf_all <- gf_outside_cluster %>%
  group_by(YEAR, TIME_STEP_KEY, STAT_AREA, PROCESSING_SECTOR) %>%
  reframe(
    tons_gf_wk = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE),
    mean_psc_wk = mean(PSC_VALUE, na.rm = TRUE),
    mean_gf_tons_wk = mean(GF_TOTAL_CATCH_MT, na.rm = TRUE)) 

sum.stat.timeseries.gf_all$mean_psc_rate_wk <- ifelse(
  sum.stat.timeseries.gf_all$mean_gf_tons_wk > 0,
  sum.stat.timeseries.gf_all$mean_psc_wk / sum.stat.timeseries.gf_all$mean_gf_tons_wk,
  0
)


# Proportions by YEAR and STAT_AREA  ---
gf.outside.stat <- gf_outside_cluster %>%
  group_by(YEAR, TIME_STEP_KEY, STAT_AREA, PROCESSING_SECTOR) %>%
  reframe(tons_gf_annual_stat = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE))

gf.outside.annually <- gf_outside_cluster %>%
  group_by(YEAR, TIME_STEP_KEY, PROCESSING_SECTOR) %>%
  reframe(tons_gf_annual = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE))

sum.stat.timeseries.gf_all2 <- dplyr::left_join(sum.stat.timeseries.gf_all, gf.outside.annually, by = c("YEAR", "TIME_STEP_KEY", "PROCESSING_SECTOR"))
sum.stat.timeseries.gf_all3 <- dplyr::left_join(sum.stat.timeseries.gf_all2, gf.outside.stat, by = c("YEAR","TIME_STEP_KEY", "STAT_AREA", "PROCESSING_SECTOR"))

sum.stat.timeseries.gf_all3$proportion_outside_wk <- ifelse(
  sum.stat.timeseries.gf_all3$tons_gf_annual > 0,
  sum.stat.timeseries.gf_all3$tons_gf_annual_stat / sum.stat.timeseries.gf_all3$tons_gf_annual,
  0
)

sum.stat.timeseries.gf_all3 <- as.data.frame(sum.stat.timeseries.gf_all3)

```

## 9. Plot Weekly Proportions

This plot shows the proportions of groundfish catch in the stat areas outside of the closure within the buffer. These proportions will be what displaced catch is applied to, under the assumption that the fishery will meet the same overall total catch.

```{r}

#| label: fig-props
#| fig-cap: "Observed groundfish redistribution allocation proportions."
#| fig-width: 12
#| fig-height: 8

# Proportion join
sum.stat.timeseries.gf_all_shp <- dplyr::left_join(stat.areas.timeseries, sum.stat.timeseries.gf_all3, by=c("YEAR", "STAT_AREA"))
sum.stat.timeseries.gf_all_shp <- terra::na.omit(sum.stat.timeseries.gf_all_shp, "PROCESSING_SECTOR")

propdat <- sum.stat.timeseries.gf_all_shp
propdat$label <- round(propdat$proportion_outside_wk, 2)

ggplot() +
  tidyterra::geom_spatvector(data=subset(propdat, propdat$YEAR==2015, propdat$proportion_outside_wk>0), aes(fill=proportion_outside_wk), alpha = 0.8, col=NA, na.rm=TRUE) +
  scale_fill_distiller(palette = "GnBu", name = "Proportion of GF Catch") +
  tidyterra::geom_spatvector(data = closure2, aes(color = "Closure"), 
                             fill = NA, alpha = 0.5, lwd = 0.6) +
  scale_color_manual(values = c("Closure" = "red"), name = NULL) +
  facet_wrap(~TIME_STEP_KEY, ncol=10) +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5), axis.title = element_blank())

```

## 10. Calculate Weekly and Annual Summary Statistics

This step applies the displacement and new PSC calculations.

```{r}

#| label: summary-stats
#| echo: false

# Join catch from inside area
total.join <- dplyr::left_join(sum.stat.timeseries.gf_all3,
                               gf_inside_cluster2 %>% dplyr::select(YEAR, TIME_STEP_KEY, PROCESSING_SECTOR, tons_gf_inside_wk),
                               by=c("YEAR", "TIME_STEP_KEY",  "PROCESSING_SECTOR"))

# NAs to 0s to remove weeks where no catch happened inside the closed area
total.join$tons_gf_inside_wk[is.na(total.join$tons_gf_inside_wk)] <- 0

# Multiply weekly displaced catch by the outside proportions
total.join$displaced_gf_catch_wk <- total.join$tons_gf_inside_wk * total.join$proportion_outside_wk

# Multiply PSC rate by new gf catch
total.join$new_PSC_wk <- total.join$displaced_gf_catch_wk * total.join$mean_psc_rate_wk


# Summarize by week and sector
# Note: sum(unique(...)) prevents multi-sector records from over-inflating the baseline displacement total
total.join2 <- total.join %>% 
  group_by(YEAR, TIME_STEP_KEY, PROCESSING_SECTOR) %>%
  reframe(
    tons_gf_inside_wk = sum(unique(tons_gf_inside_wk)), 
    displaced_gf_catch_wk = sum(displaced_gf_catch_wk),
    new_PSC_wk = sum(new_PSC_wk),
    mean_psc_rate_wk = mean(mean_psc_rate_wk))

# Join the status quo inside psc numbers
total.join3 <- dplyr::left_join(total.join2,
                                gf_inside_cluster2 %>% dplyr::select(YEAR, TIME_STEP_KEY, PROCESSING_SECTOR, psc_measure_inside_wk),
                                by=c("YEAR", "TIME_STEP_KEY", "PROCESSING_SECTOR"))
# NAs to 0s
total.join3$psc_measure_inside_wk[is.na(total.join3$psc_measure_inside_wk)] <- 0

# Subtract the PSC caught inside the area
total.join3$net_change_PSC_wk <- total.join3$new_PSC_wk - total.join3$psc_measure_inside_wk

total.join.plot <- total.join3


# Sum est changes for annual numbers
total.join.plot.annual <- total.join.plot %>%
  group_by(YEAR) %>%
  reframe(
    est_change_fleet = sum(net_change_PSC_wk)
  )

```


## 11. Bootstrap

To provide a measure of uncertainty, we are bootstrapping 1,000 alternative scenarios by randomly shuffling the PSC rates where fishing is redistributed. We calculated a 95% Confidence Interval using the middle 95% of all simulated outcomes (between 2.5th and 97.5th percentiles).

```{r}

# ==============================================================================
# SPATIAL BYCATCH RATE BOOTSTRAP 
# ==============================================================================
#| label: spatial-rate-bootstrap

set.seed(345)
n_iterations <- 1000
time_steps   <- unique(total.join$TIME_STEP_KEY)

# Matrix to store results (Rows = Iterations, Cols = Time Steps)
boot_matrix <- matrix(NA, nrow = n_iterations, ncol = length(time_steps))
colnames(boot_matrix) <- time_steps

# FIX: Explicitly pull psc_measure_inside_wk from gf_inside_cluster2 during setup
df_bs_base <- total.join %>%
  dplyr::left_join(
    gf_inside_cluster2 %>% dplyr::select(YEAR, TIME_STEP_KEY, PROCESSING_SECTOR, psc_measure_inside_wk),
    by = c("YEAR", "TIME_STEP_KEY", "PROCESSING_SECTOR")
  ) %>%
  dplyr::mutate(
    psc_measure_inside_wk = tidyr::replace_na(psc_measure_inside_wk, 0),
    displaced_gf_catch_wk = tons_gf_inside_wk * proportion_outside_wk
  )

for(i in 1:n_iterations) {
  
  # Vector to temporarily store the randomized PSC calculations for this iteration
  new_psc_iteration <- numeric(nrow(df_bs_base))
  
  for(ts in time_steps) {
    # Isolate the row indices belonging to this specific week/month step
    ts_indices <- which(df_bs_base$TIME_STEP_KEY == ts)
    
    # Extract the pool of observed rates for this specific time frame
    psc_pool <- df_bs_base$mean_psc_rate_wk[ts_indices]
    
    # If the pool is valid, randomly shuffle the rates across the active areas
    if(length(psc_pool) > 0) {
      sampled_rates <- sample(psc_pool, size = length(ts_indices), replace = TRUE)
      new_psc_iteration[ts_indices] <- df_bs_base$displaced_gf_catch_wk[ts_indices] * sampled_rates
    }
  }
  
  # Summarize the net impact for this specific shuffle loop
  iter_summary <- df_bs_base %>%
    dplyr::mutate(new_psc_boot = new_psc_iteration) %>%
    dplyr::group_by(TIME_STEP_KEY) %>%
    dplyr::summarise(
      net_change = sum(new_psc_boot, na.rm = TRUE) - max(psc_measure_inside_wk, na.rm = TRUE),
      .groups = "drop"
    )
  
  # Map iteration calculations into our tracking matrix
  for(j in seq_along(time_steps)) {
    boot_matrix[i, j] <- iter_summary$net_change[iter_summary$TIME_STEP_KEY == time_steps[j]]
  }
}

# Transform the operational matrix into a long format data frame
boot_long <- as.data.frame(boot_matrix) %>%
  dplyr::mutate(iteration = 1:n()) %>%
  tidyr::pivot_longer(-iteration, names_to = "TIME_STEP_KEY", values_to = "net_change")

# Link calendar YEARS back to our time-step keys for clear annual grouping
year_mapping <- total.join %>% 
  dplyr::select(TIME_STEP_KEY, YEAR) %>% 
  dplyr::distinct()

boot_annual_dist <- boot_long %>%
  dplyr::left_join(year_mapping, by = "TIME_STEP_KEY") %>%
  dplyr::group_by(iteration, YEAR) %>%
  dplyr::summarise(annual_psc_change = sum(net_change, na.rm = TRUE), .groups = "drop")

# Calculate the final formal 95% Confidence Bounds (2.5th and 97.5th percentiles)
boot_annual_ci <- boot_annual_dist %>%
  dplyr::group_by(YEAR) %>%
  dplyr::summarise(
    lower_change = quantile(annual_psc_change, 0.025, na.rm = TRUE),
    upper_change = quantile(annual_psc_change, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

```

## 12. Evaluate Impacts

Compares the estimated PSC change to the observed PSC (status quo) each year. For this example, summarized by year.

```{r}

#| label: fig-status-quo
#| fig-width: 12
#| fig-height: 5

# --- total by year (only estimated change) with status quo numbers --- #
total.join.plot.annual2 <- dplyr::left_join(all_psc_status_quo_annual, total.join.plot.annual, by = c("YEAR"))
total.join.plot.annual2$est_change_fleet[is.na(total.join.plot.annual2$est_change_fleet)] <- 0
total.join.plot.annual2$net_new_PSC <- total.join.plot.annual2$sum_psc_status_quo + total.join.plot.annual2$est_change_fleet
total.join.plot.annual2$LABEL <- paste0("SPECIES")
colnames(total.join.plot.annual2) <- c("YEAR", "psc_statusquo", "psc_change", "net_new_PSC", "LABEL")

# STEP 1: Join and calculate the upper/lower bounds BEFORE pivoting the data
total.join.plot.annual2 <- total.join.plot.annual2 %>%
  dplyr::left_join(boot_annual_ci, by = "YEAR") %>%
  dplyr::mutate(
    boot_ymin = psc_statusquo + lower_change,
    boot_ymax = psc_statusquo + upper_change
  )

# STEP 2: Pivot into long format (keeping our new boot columns intact)
total.join.plot.annual3 <- total.join.plot.annual2 %>% pivot_longer(
  cols = c("net_new_PSC", "psc_statusquo"),
  names_to = "psc_type",
  names_prefix = "psc_",
  values_to = "psc_amount",
  values_drop_na = TRUE
)

# STEP 3: Assign the error bar values exclusively to the net_new_PSC bar rows
total.join.plot.annual3 <- total.join.plot.annual3 %>%
  dplyr::mutate(
    ymin = ifelse(psc_type == "net_new_PSC", boot_ymin, NA_real_),
    ymax = ifelse(psc_type == "net_new_PSC", boot_ymax, NA_real_)
  )

# #plot (Updated x to factor(YEAR) to align real historical timelines cleanly)
ggplot(total.join.plot.annual3, aes(x = factor(YEAR), y = psc_amount, fill = psc_type)) +
  ggtitle(paste0(spp_title, ": ", title_name)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7), width = 0.7) +
  geom_errorbar(
    aes(ymin = ymin, ymax = ymax, group = psc_type),
    position = position_dodge(width = 0.7),
    width = 0.2,
    color = "black",
    linewidth = 0.7,
    na.rm = TRUE
  ) +
  theme_bw() +
  labs(x = "Year", y = "Estimated Annual PSC (# Fish / Wt)") +
  scale_fill_brewer(palette = "Paired", name = "Closure Scenario", labels = c("RKCSSA", "Status Quo")) +
  theme(plot.title = element_text(hjust = 0.5))

```

## 13. Summary Statistics

```{r}

#| label: final-table
#| fig-width: 8
#| fig-height: 5

# Final summary table
final.summary.table <- total.join.plot.annual2 %>% dplyr::select(YEAR, psc_statusquo, net_new_PSC, psc_change)
final.summary.table <- as.data.frame(final.summary.table)
colnames(final.summary.table) <- c("year", "PSC_status_quo", "PSC_closure", "Net_Change")

# Safe percentage calculation handling potential zero-activity years
final.summary.table$percent_change <- ifelse(
  final.summary.table$PSC_status_quo > 0,
  (final.summary.table$Net_Change / final.summary.table$PSC_status_quo) * 100,
  0
)

print(final.summary.table)

```

## 14. Validate Movement - 1-Year Lag

To validate the accuracy of the model's displacement predictions, we use a 1-year time lag to compare the proportion of groundfish catch in the outside stat_areas after a closure to the actual observations the following year under actual closures in 2022 and 2023. For this example, we show a correlation of 0.66 in the first year of the closure, and 0.89 in the second. This supports the use of historical catch proportions as a proxy for predicting where vessels with go when faced with a closure.

```{r}

# ==============================================================================
# LAGGED PREDICTOR BASELINE (Prior Years: 2021 & 2022)
# ==============================================================================
#| label: predictive-baseline-lagged

prop_df_clean <- as.data.frame(propdat)

# Extract the prior years' distributions using your exact structural layout
annual_spatial_distribution_lagged <- prop_df_clean %>%
  dplyr::filter(YEAR %in% c(2021, 2022)) %>% 
  dplyr::group_by(YEAR, STAT_AREA, PROCESSING_SECTOR) %>% 
  dplyr::summarise(
    annual_mean_stat_prop_prior = mean(proportion_outside_wk, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # BRIDGE STEP: Advance the year value by +1 so it aligns with the target validation year
  dplyr::mutate(
    VALIDATION_YEAR_MATCH = YEAR + 1
  )

# ==============================================================================
# ACTUAL OBSERVED (Validation Years: 2022 & 2023)
# ==============================================================================
#| label: validation-footprint

# Isolate your validation years using the exact same structure
annual_spatial_distribution_validation <- prop_df_clean %>%
  dplyr::filter(YEAR %in% c(2022, 2023)) %>% 
  dplyr::group_by(YEAR, STAT_AREA, PROCESSING_SECTOR) %>%  
  dplyr::summarise(
    annual_mean_stat_prop_actual = mean(proportion_outside_wk, na.rm = TRUE),
    .groups = "drop"
  )

# ==============================================================================
# DIRECT VALIDATION JOIN & PLOT (Prior-Year Lagged Evaluation)
# ==============================================================================
#| label: validation-diagnostics

# Join them by aligning the actual validation year with the lagged predictor's match key
validation_master_table <- annual_spatial_distribution_validation %>%
  dplyr::inner_join(
    annual_spatial_distribution_lagged, 
    by = c("YEAR" = "VALIDATION_YEAR_MATCH", "STAT_AREA", "PROCESSING_SECTOR")
  )

# Calculate simple accuracy metrics comparing actual distributions to the prior year's footprint
validation_metrics <- validation_master_table %>%
  dplyr::group_by(YEAR, PROCESSING_SECTOR) %>%
  dplyr::summarise(
    RMSE = sqrt(mean((annual_mean_stat_prop_actual - annual_mean_stat_prop_prior)^2, na.rm = TRUE)),
    Correlation = cor(annual_mean_stat_prop_actual, annual_mean_stat_prop_prior, use = "complete.obs"),
    .groups = "drop"
  )

print(as.data.frame(validation_metrics))

# Plot the short-term memory model fit
ggplot(validation_master_table, aes(x = annual_mean_stat_prop_prior, y = annual_mean_stat_prop_actual, color = factor(YEAR))) +
  geom_point(size = 3.5, alpha = 0.8) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray40") +
  theme_bw() +
  labs(
    title = paste0("Prior-Year Lagged Validation Fit: ", title_name),
    x = "Prior Year Expected Proportions (2021 & 2022)",
    y = "Actual Observed Proportions (2022 & 2023)",
    color = "Closure Year"
  ) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

```

## 15. How did our PSC estimates do?

To validate the accuracy of our PSC estimates under the closure, we used a historical baseline (2009-2021) of total annual PSC rates (Z_hist = PSC_hist / GF_hist) to compare to the predicted (Z_pred) and actual (Z_obs) PSC rates under the groundfish catch observed in each closure year (GF_obs). PSC catch = PSC Rate * GF Catch.

```{r}

#| label: global-volume-triple-check
#| fig-width: 8
#| fig-height: 5

# 1. Calculate the historical aggregate baseline rate (2009-2021 Outside)
historical_baseline_rate <- ledger_df_clean %>%
  dplyr::filter(YEAR %in% c(2009:2021)) %>%
  dplyr::summarise(
    total_hist_psc = sum(PSC_VALUE, na.rm = TRUE),
    total_hist_gf  = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    mean_baseline_rate = ifelse(total_hist_gf > 0, total_hist_psc / total_hist_gf, 0)
  ) %>%
  dplyr::pull(mean_baseline_rate)

# 2. Calculate the standard deviation of the historical annual rates (2009-2021 Outside)
annual_rates_hist <- ledger_df_clean %>%
  dplyr::filter(YEAR %in% c(2009:2021)) %>%
  dplyr::group_by(YEAR) %>%
  dplyr::summarise(
    annual_psc = sum(PSC_VALUE, na.rm = TRUE),
    annual_gf  = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    annual_rate = ifelse(annual_gf > 0, annual_psc / annual_gf, 0)
  )

historical_baseline_sd <- sd(annual_rates_hist$annual_rate, na.rm = TRUE)
if (is.na(historical_baseline_sd)) historical_baseline_sd <- 0

# 3. Model-Implied Post-Closure Rate Baseline (2009-2021 Simulated)
model_closure_rate <- total.join %>%
  dplyr::filter(YEAR %in% c(2009:2021)) %>%
  dplyr::summarise(
    total_model_psc = sum(new_PSC_wk, na.rm = TRUE),
    total_model_gf  = sum(tons_gf_annual_stat, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    predicted_closure_rate = ifelse(total_model_gf > 0, total_model_psc / total_model_gf, 0)
  ) %>%
  dplyr::pull(predicted_closure_rate)

# 4. Scale tracks against actual 2022/2023 landings
triple_volume_comparison <- ledger_df_clean %>%
  dplyr::filter(YEAR %in% c(2022, 2023)) %>%
  dplyr::group_by(YEAR, PROCESSING_SECTOR) %>%
  dplyr::summarise(
    actual_tonnage_landed = sum(GF_TOTAL_CATCH_MT, na.rm = TRUE),
    Actual_Observed_PSC_T  = sum(PSC_VALUE, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    Unconstrained_Historical_PSC_Z_hist    = actual_tonnage_landed * historical_baseline_rate,
    Unconstrained_Historical_PSC_Z_hist_sd = actual_tonnage_landed * historical_baseline_sd,
    Model_Predicted_PSC_Z_model            = actual_tonnage_landed * model_closure_rate
  )

# Pivot data into long format for clear multi-bar grouping
plot_triple_data <- triple_volume_comparison %>%
  dplyr::select(YEAR, Unconstrained_Historical_PSC_Z_hist, Model_Predicted_PSC_Z_model, Actual_Observed_PSC_T) %>%
  tidyr::pivot_longer(
    cols = c("Unconstrained_Historical_PSC_Z_hist", "Model_Predicted_PSC_Z_model", "Actual_Observed_PSC_T"),
    names_to = "Scenario_Type",
    values_to = "PSC_Count"
  )

# Join the error metrics back to the long plotting table
plot_triple_data <- plot_triple_data %>%
  dplyr::left_join(triple_volume_comparison %>% dplyr::select(YEAR, Unconstrained_Historical_PSC_Z_hist_sd), by = "YEAR") %>%
  dplyr::mutate(
    # Apply variance limits ONLY to the Historical bar, leave others as NA
    ymin = ifelse(Scenario_Type == "Unconstrained_Historical_PSC_Z_hist", pmax(0, PSC_Count - Unconstrained_Historical_PSC_Z_hist_sd), NA_real_),
    ymax = ifelse(Scenario_Type == "Unconstrained_Historical_PSC_Z_hist", PSC_Count + Unconstrained_Historical_PSC_Z_hist_sd, NA_real_)
  )

# Set up clean factor levels to dictate left-to-right bar ordering
plot_triple_data$Scenario_Type <- factor(
  plot_triple_data$Scenario_Type,
  levels = c("Unconstrained_Historical_PSC_Z_hist", "Model_Predicted_PSC_Z_model", "Actual_Observed_PSC_T")
)

# Generate the Plot
ggplot(plot_triple_data, aes(x = factor(YEAR), y = PSC_Count, fill = Scenario_Type)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.75), alpha = 0.9, width = 0.75) +
  geom_errorbar(
    aes(ymin = ymin, ymax = ymax, group = Scenario_Type), 
    position = position_dodge(width = 0.75),             
    width = 0.25,
    color = "black",
    linewidth = 0.8,
    na.rm = TRUE
  ) +
  scale_fill_manual(
    values = c(
      "Unconstrained_Historical_PSC_Z_hist" = "#757575", # Dark Grey
      "Model_Predicted_PSC_Z_model"         = "#e41a1c", # Red
      "Actual_Observed_PSC_T"               = "#377eb8"  # Blue
    ),
    labels = c(
      "Historical Baseline (Z_hist) ± SD", 
      "Model-Predicted Closure (Z_model)", 
      "Actual Observed Reality (T)"
    )
  ) +
  theme_bw() +
  labs(
    title = paste0("Closure Validation: ", title_name),
    x = "Emergency Closure Year",
    y = paste0("Total PSC (", params$psc_spp, " Count)"),
    fill = "Scenario Track"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    plot.subtitle = element_text(face = "italic", size = 9, hjust = 0.5),
    legend.position = "bottom",
    legend.direction = "vertical",
    legend.box = "horizontal"
  )

```


